---
title: "Modern concepts and methods in Macroecology and Biogeography"
subtitle: "Day 1: Basics and spatial data in R"
author: "Dylan Craven & Patrick Weigelt"
date: "April 16, 2019"
output: 
  html_document: 
    theme: flatly
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

&nbsp;
_[back to overview](index.html)_
&nbsp;
&nbsp;

## 1. Preparation

### a) Prepare your system for exchanging data
Set the format your computer uses to display numbers to 
English format for exchanging data with others:

* Decimal symbol: " . "  
* Digit grouping symbol: " , "  
* List seperator: " , "  
&nbsp;

### b) Install software

* Get QGIS from <https://www.qgis.org/>
* Get R from <http://www.r-project.org/>
* Get RStudio from <https://www.rstudio.com/>

### c) Create a new project in RStudio

_Do this only once for the entire practical part of the class and once again for your project in the end but not for each session separately_
&nbsp;

* Open RStudio  
* File | New Project
* Select *either* 'New Directory' or 'Existing Directory'   
  this will be the working folder for our course (on your computer)
* Locate the project on your computer in a place 
  (e.g. '~Desktop/Courses/Macroecology)  
* Name your project  
* Within your project's folder, create three sub-folders: 
    + data
    + code
    + figures
* Create a new empty R script: File | New File | R script
* Type: # Day 1: getting started
* Save file (in ~/code): "0_Code_Day1.R" 
&nbsp;
* Start coding!


## 2. Getting started
**Using R as a calulator**

```{r, eval=FALSE}
# use the console to do simple calculations
2 + 2
2 * 2
8 / 2
6 - 2
sqrt(16) # sqrt() is a function
16^(1/2)
2^2
1/2 ^ (-2)
log10(10000)
```
What is the answer of each of the operations?  
&nbsp;


**Assign a value to an object**

_Numbers_
```{r}
a <- 4
b <- sqrt(16)
number_3 <- 2 * 2
```

```{r}
a
number_3
a*b
number_3 / b
```
&nbsp;


_Character strings_
```{r}
text_a <- "Hello"
text_a

text_a <- "Good bye" # overwrite the object
text_a
```
&nbsp;


In R Studio the objects are summed up at the top-right corner. For seeing the stored objects use:
```{r}
ls()
```
&nbsp;


## 3. Getting help in R

`?function.name`
`??keyword`

```{r, eval=FALSE}
?sqrt # will open help page in Help tab of RStudio or browser
??squareroot
```
&nbsp;


## 4. Data structures and indexing

### a) Numbers \& characters
_single values_

```{r}
mode(a)
str(a)
```

```{r}
mode(text_a)
str(text_a)
```
&nbsp;


### b) Vectors
_sequences of values of the same mode_

**Create a vector of numbers**
```{r}
vector_1 <- c(1,5,6,3.4,10,11)
vector_1
```
&nbsp;

**Indexing and modifying a vector**
```{r}
vector_1[3]
vector_1[3] <- 5
vector_1[3]
```
&nbsp;

**Creating a sequence of numbers**
```{r}
vector_2 <- c(1:6)
vector_2
vector_2 <- seq(1,12,by=2) # overwriting
vector_2
```

```{r}
vector_3 <- rep(c(1,2),3) # repeats c(1,2) three times
vector_3
```
&nbsp;

**Create a vector of character strings**
```{r}
vector_4 <- c("blue","red","green","yellow","grey","black")
vector_4
vector_4[4]
```

```{r}
mode(vector_4)
length(vector_2)
mode(vector_2)
```
&nbsp;

You can do calculations with numeric vectors just like with values
```{r}
mean(vector_2) # function for calculating the mean
sd(vector_2)   # standard deviation
```

```{r}
vector_2
vector_2 ^ 2

vector_5 <- vector_1 + vector_2
vector_5
```
&nbsp;

**Plotting vectors**
```{r}
plot(vector_2,vector_1)
```


### c) Arrays and matrices
_tables of values of the same mode with n dimensions (matrix = 2 dimensions)_
```{r}
matrix_1 <- matrix(NA, nrow = 6, ncol = 6)
matrix_1

matrix_1 <- matrix(1:36, nrow = 6, ncol = 6) # Filling a matrix with a sequence of values
matrix_1

matrix_2 <- matrix(1:(6*6), nrow = 6, ncol = 6, byrow = TRUE)
matrix_2
```
&nbsp;

**Indexing a matrix**
```{r}
matrix_2[3,5] # row,column
matrix_2[3,]  # select a row
matrix_2[,5]  # select a column
```
&nbsp;

You can do calculations with numeric matrices just like with values and vectors
```{r}
matrix_1 + 1
matrix_1 ^ 2
```

```{r}
matrix_1 + matrix_2 # matrix by matrix
matrix_1 * vector_2 # matrix by vector; vector of length 6 is recycled!
```

```{r}
mean(matrix_1)     # mean of all values in matrix
colMeans(matrix_1) # mean values per column
```
&nbsp;


### d) data.frames 
_tables of values of one mode per column_

**Create a data.frame**
```{r}
df_1 <- data.frame(ID = vector_2, number_1 = vector_1, colour = vector_4)
df_1
```


```{r}
str(df_1) 
```
"colour" is a factor; makes sense if certain values or levels are repeated

&nbsp;

**Indexing**
```{r}
df_1[,3]        # by column number (also works for rows; [2,])
df_1[,"colour"] # calling by column name
df_1$colour     # calling by column name using $
```

```{r}
levels(df_1[,3])
df_1[,3] <- as.character(df_1[,3]) # change column to character vector
str(df_1)
```
colour is a character vector
&nbsp;

```{r}
plot(df_1$ID,df_1$number_1, col=df_1$colour)
```

## 5. data.frame exercise  

1) Imagine a field study with 5 plots (1,2,3,...) in each of 3 vegetation types (forest, shrubland, grassland).  
2) Create a data.frame with 1 row per plot and 2 columns for "vegetation_type" and "plot".  
3) Every second day of a month a student visits one of the plots. Add a third column called "day" indicating when the student visits each plot: 1, 3, 5, ...

```{r, echo=FALSE}
plots <- data.frame(vegetation_type = rep(c("forest", "shrubland", "grassland"),each=5), plot = rep(1:5,3))
plots$day <- seq(1,29,by=2)
knitr::kable(plots, caption="Table 1: Dataframe exercise")
```
&nbsp;


## 6. Loading and visualising environmental raster layers  

**Download the following raster layers**  
_GMTED - Digital elevation data <https://lta.cr.usgs.gov/GMTED2010>_  
[Elevation above sea level (mn30_grd.zip)](http://edcintl.cr.usgs.gov/downloads/sciweb1/shared/topo/downloads/GMTED/Grid_ZipFiles/mn30_grd.zip)

_Chelsa - Climate data <http://chelsa-climate.org/>_  
[Annual mean temperature (CHELSA_bio10_01_land.7z)](http://www.wsl.ch/lud/chelsa/data/bioclim/integer/CHELSA_bio10_01_land.7z)  
[Mean annual precipitation (CHELSA_bio10_12_land.7z)](http://www.wsl.ch/lud/chelsa/data/bioclim/integer/CHELSA_bio10_12_land.7z)  
&nbsp;

**Extract tand copy the rsaters into your R-project folder ../data/**  
```{r, echo=FALSE}
getwd()
```

**Exercise: Open QGIS and add the raster layers to a project in QGIS**  
&nbsp;

**Load raster library**  
```{r}
# install.packages("raster") # install the raster package in case not done yet
library(raster, quietly = TRUE)
```
```{r, echo=FALSE}
if (file.exists("D:/rastertemp")){rasterOptions(tmpdir="D:/rastertemp/")} # raster saves large temporary files in your user operating systems` user folder. You might want to change this if you're running out of disk space
```
&nbsp;

**Import global digital elevation model and plot it**  
```{r}
elev <- raster("data/mn30_grd")
par(mfrow=c(1,1))
plot(elev)
```
&nbsp;

**Import climate data from Chelsa and plot it**  
```{r}
bio1 <- raster("data/CHELSA_bio10_1.tif")
bio12 <- raster("data/CHELSA_bio10_12.tif")
```
```{r}
plot(bio1)  # Temperature (*10 °C)
plot(bio12) # Precipitation (mm)
```
&nbsp;

**Have a look at characteristics of the raster layers**  
`str(bio1) # long output; try it out`

```{r}
crs(bio1)
bio1@crs
bio1@ncols
bio1@nrows
area(bio1)
```
30 arc seconds resolution = ~1km² at the equator
&nbsp;

**You can do calculations with the raster layers**  
```{r}
bio1 <- bio1/10      # Careful! this is computer intense and might take a long time
plot(bio1)           # Temperature (°C)
points(9.916,51.541, pch=4) # Göttingen
text(9.916,51.541, "You are here", pos = 4)
```




