---
title: "Modern concepts and methods in Macroecology and Biogeography"
subtitle: "Day 1: Basics and spatial data in R"
author: "Patrick Weigelt, Pierre Denelle & Holger Kreft"
date: "April 10th, 2024"
output: 
  html_document: 
    theme: flatly
    toc: yes
editor_options:
 chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c("top", "right"), color = "black")
```

&nbsp;
_[back to overview](index.html)_
&nbsp;
&nbsp;

# 1. Preparation

## a) Prepare your system for exchanging data
Set the format your computer uses to display numbers to 
English format for exchanging data with others:

* Decimal symbol: " . "  
* Digit grouping symbol: " , "  
* List separator: " , "  
&nbsp;

## b) Install software

* Get QGIS from <https://www.qgis.org/>
* Get R from <https://cran.r-project.org/>
* Get RStudio from <https://www.rstudio.com/products/rstudio/download/>

## c) Create a new project in RStudio

_Do this only once for the entire practical part of the class and once again for your project in the end but not for each session separately_
&nbsp;

* Open RStudio  
* File | New Project
* Select *either* 'New Directory' or 'Existing Directory'   
  this will be the working folder for our course (on your computer)
* Navigate to the prospective project folder on your computer 
  (e.g. '~Documents/Courses/Macroecology)  
* Name your project  
* Within your project's folder, create three sub-folders: 
    + data
    + code
    + figures
* Create a new empty R script: File | New File | R script
* Type: # Day 1: getting started
* Save file (in ~/code): "0_Code_Day1.R" 
&nbsp;
* Start coding!


If you want to visualize this tutorial in the viewer inside RStudio (to save space on your screen) run the following chunk of code:
```{r viewer, eval=FALSE}
install.packages("rstudioapi") # install an R-package required for this step
dir <- tempfile()
dir.create(dir)
download.file("https://gift.uni-goettingen.de/mcmmb/index.html", destfile = file.path(dir, "index.html"))
download.file("https://gift.uni-goettingen.de/mcmmb/Day1.html", destfile = file.path(dir, "Day1.html"))
htmlFile <- file.path(dir, "Day1.html")
rstudioapi::viewer(htmlFile)
```
Now you can conveniently copy code from the viewer into your script.
&nbsp;

# 2. Getting started
**Using R as a calculator**

```{r, eval=FALSE}
# use the console to do simple calculations
2 + 2
2 * 2
8 / 2
6 - 2
sqrt(16) # sqrt() is a function
16^(1/2)
2^2
1/2 ^ (-2)
log10(10000)
```
What is the answer of each of the operations?  
&nbsp;

**Assign a value to an object**

_Numbers_
```{r}
a <- 4
b <- sqrt(16)
number_3 <- 2 * 2
```

```{r}
a
number_3
a*b
number_3 / b
```
&nbsp;

_Character strings_
```{r}
text_a <- "Hello"
text_a

text_a <- "Good bye" # overwrite the object
text_a
```
&nbsp;

In R Studio current objects are listed at the top-right corner. For seeing the stored objects in the console use:
```{r}
ls()
```
&nbsp;

# 3. Getting help in R

`?function.name`
`??keyword`

```{r, eval=FALSE}
?sqrt # will open help page in Help tab of RStudio or browser
??squareroot
??`install package` # Or use google
```
&nbsp;

# 4. Data structures and indexing
## a) Numbers \& characters
_single values_

```{r}
mode(a)
str(a)
```

```{r}
mode(text_a)
str(text_a)
```
&nbsp;

## b) Vectors
_sequences of values of the same mode_

**Create a vector of numbers**
```{r}
vector_1 <- c(1,5,6,3.4,10,11)
vector_1
```
&nbsp;

**Indexing and modifying a vector**
```{r}
vector_1[3]
vector_1[3] <- 5
vector_1[3]
```
&nbsp;

**Creating a sequence of numbers**
```{r}
vector_2 <- c(1:6)
vector_2
vector_2 <- seq(1, 12, by=2) # overwriting
vector_2
vector_2 <- seq(1, 11, length.out = 6) # overwriting
vector_2
```

```{r}
vector_3 <- rep(c(1,2),3) # repeats c(1,2) three times
vector_3
vector_3 <- rep(c(1,2),each=3) # repeats each element of c(1,2) three times
vector_3
```
&nbsp;

**Create a vector of character strings**
```{r}
vector_4 <- c("blue","red","green","yellow","grey","black")
vector_4
vector_4[4]
```

```{r}
mode(vector_4)
length(vector_2)
mode(vector_2)
```
&nbsp;

You can do calculations with numeric vectors just like with values
```{r}
vector_2
vector_2^2

vector_5 <- vector_1 + vector_2
vector_5
```
&nbsp;

You can apply functions to vectors
```{r}
mean(vector_2) # function for calculating the mean
sd(vector_2)   # standard deviation
```

_Functions require a certain input structure like here a vector_  
```{r}
mean(c(3,6))
mean(3,6) # actually means mean(x=3, trim=6, na.rm = FALSE)
#?mean
```
&nbsp;

**Plotting vectors**
```{r}
plot(vector_2,vector_1)
```

There are two major ways of plotting in R: Base R like above and [ggplot2](https://www.cedricscherer.com/2019/08/05/a-ggplot2-tutorial-for-beautiful-plotting-in-r/) which we will touch upon throughout the tutorials.
&nbsp;

## c) Arrays and matrices
_tables of values of the same mode with n dimensions (matrix = 2 dimensions)_
```{r}
matrix_1 <- matrix(NA, nrow = 6, ncol = 6)
matrix_1

matrix_1 <- matrix(1:36, nrow = 6, ncol = 6) # Filling a matrix with a sequence of values
matrix_1

matrix_2 <- matrix(1:(6*6), nrow = 6, ncol = 6, byrow = TRUE)
matrix_2
```
&nbsp;

**Indexing a matrix**
```{r}
matrix_2[3,5] # row,column
matrix_2[3,]  # select a row
matrix_2[,5]  # select a column
```
&nbsp;

You can do element-wise calculations with numeric matrices just like with values and vectors
```{r}
matrix_1 + 1
matrix_1 ^ 2
```

```{r}
matrix_1 + matrix_2 # matrix by matrix
matrix_1 * vector_2 # matrix by vector; vector of length 6 is recycled! In this case each column is multiplied by the vector because they have the same length
```

```{r}
mean(matrix_1)     # mean of all values in matrix
colMeans(matrix_1) # mean values per column
```
You can also do proper matrix calculations like Matrix multiplication or cross-products. For example, see [this](https://www.statmethods.net/advstats/matrix.html).

&nbsp;

## d) data.frames 
_tables of values of one mode per column_

**Create a data.frame**
```{r}
df_1 <- data.frame(ID = vector_2, number_1 = vector_1, colour = vector_4)
df_1
```

```{r}
str(df_1) 
```
"colour" is a character vector

&nbsp;

**Indexing**
```{r}
df_1[,3]        # by column number (also works for rows; [2,])
df_1[,"colour"] # calling by column name
df_1$colour     # calling by column name using $
```

```{r}
df_1[,3] <- as.factor(df_1[,3]) # change column to factor
levels(df_1[,3])
str(df_1)
```
Now "colour" is a factor, which is smaller when stored and faster to handle.  
&nbsp;
&nbsp;

Be careful when converting factors to numeric vectors or columns!  
```{r}
numeric_factor <- as.factor(df_1$number_1)
numeric_factor
as.numeric(numeric_factor)
as.numeric(as.character(numeric_factor))
```
&nbsp;

# 5. data.frame exercise  

1) Imagine a field study with 5 plots (1,2,3,...) in each of 3 vegetation types (forest, shrubland, grassland).  
2) Create a data.frame with 1 row per plot and 2 columns for "vegetation_type" and "plot".  
3) Every second day of a month a student visits one of the plots. Add a third column called "day" indicating when the student visits each plot: 1, 3, 5, ...

```{r, echo=FALSE}
plots <- data.frame(vegetation_type = rep(c("forest", "shrubland", "grassland"),each=5), plot = rep(1:5,3))
plots$day <- seq(1,29,by=2)
knitr::kable(plots, caption="Table 1: Dataframe exercise")
```
&nbsp;

# 6. Plotting dataframes

In R, there are two major ways to plot your data: **base R** and **ggplot2**.  
There is no better way to plot things, it is rather a matter of taste, but it
is quite important to know how both work.  
Let's see this with an easy example.  

```{r}
library(ggplot2)

# The following R packages use color palettes are perceptually uniform, work
# with black and white, accessible to colorblind
library(RColorBrewer) # https://colorbrewer2.org/
library(viridis) # https://www.r-bloggers.com/2018/07/ggplot2-welcome-viridis/

# Example data frame
set.seed(1) # setting a random seed (to reproduce output)
ex <- data.frame(x = seq(1:100),
                 y = rnorm(100),
                 cat = c(rep("A", 50), rep("B", 50)))

par(xpd = TRUE)
plot(ex$x, ex$y,
     col = brewer.pal(n = 3, name = "Set1")[1:2][as.factor(ex$cat)],
     pch = c(16, 17)[as.factor(ex$cat)], cex = 2,
     main = "Example plot", xlab = "X axis", ylab = "Y axis",
     bty = "n")
legend("topright",
       c("Category A", "Category B"),
       inset = c(-0.1, 0),
       title = "Legend title",
       text.font = 3,
       col = c("darkblue", "firebrick3"),
       pch = c(16, 17),
       bg = "gray95")

# Same plot with ggplot2
ggplot(ex, aes(x, y)) +
  geom_point(aes(color = cat, shape = cat), size = 3) +
  scale_color_brewer("Legend title", palette = "Set1") +
  scale_shape_manual("Legend title",
                     values = c(16, 17)) +
  labs(title = "Example plot", x = "X axis", y = "Y axis") +
  theme_bw()

# With viridis color palette
ggplot(ex, aes(x, y)) +
  geom_point(aes(color = cat, shape = cat), size = 3) +
  scale_color_viridis_d("Legend title") +
  scale_shape_manual("Legend title",
                     values = c(16, 17)) +
  labs(title = "Example plot", x = "X axis", y = "Y axis") +
  theme_classic()
```

# 7. Loading and visualising environmental raster layers  

**Download the following raster layers**  
_GMTED - Digital elevation data <https://topotools.cr.usgs.gov/gmted_viewer/viewer.htm>_  
[Elevation above sea level (mn30_grd.zip)](http://edcintl.cr.usgs.gov/downloads/sciweb1/shared/topo/downloads/GMTED/Grid_ZipFiles/mn30_grd.zip)

_Chelsa - Climate data <http://chelsa-climate.org/>_  
[Annual mean temperature (CHELSA_bio10_01.tif)](https://os.zhdk.cloud.switch.ch/envicloud/chelsa/chelsa_V1/climatologies/bio/CHELSA_bio10_01.tif)  
[Mean annual precipitation (CHELSA_bio10_12.tif)](https://os.zhdk.cloud.switch.ch/envicloud/chelsa/chelsa_V1/climatologies/bio/CHELSA_bio10_12.tif)  
&nbsp;

**Extract and copy the rasters into your R-project folder ../data/**  
```{r, echo=FALSE}
getwd()
```

**Exercise: Open QGIS and add the raster layers to a project in QGIS**  
&nbsp;

**Load raster library**  
```{r, eval=FALSE}
# install.packages("terra") # install the terra package in case not done yet
library(terra)
```

```{r, echo=FALSE}
library(terra, quietly = TRUE)
```

&nbsp;

**Import global digital elevation model and plot it**  
```{r}
elev <- rast("data/mn30_grd/") + 0 # +0 to treat raster as numeric
par(mfrow=c(1,1))
plot(elev, main = "Elevation")
```
&nbsp;

**Import climate data from Chelsa and plot it**  
```{r}
bio1 <- rast("data/CHELSA_bio10_01.tif")
bio12 <- rast("data/CHELSA_bio10_12.tif")
```

```{r}
plot(bio1, main = "Temperature")  # Temperature (*10 °C)
plot(bio12, main = "Precipitation") # Precipitation (mm)
```
&nbsp;

**Have a look at characteristics of the raster layers**  
`str(bio1) # long output; try it out`

```{r}
crs(bio1)
crs(bio1)
ncol(bio1)
nrow(bio1)
cellSize(bio1)
```
30 arc seconds resolution = ~1km² at the equator
&nbsp;

**You can do calculations with the raster layers**  
```{r}
system.time(log_bio12 <- log10(bio12+1)) # Careful! this is computer intense and might take a long time
plot(log_bio12, main = "Precipitation") # Log-transformed precipitation

bio1 <- bio1/10      
plot(bio1, main = "Temperature") # Temperature (°C)
points(9.916, 51.541, pch=4) # Göttingen
text(9.916, 51.541, "You are here", pos = 4)
```
