---
title: "Modern concepts and methods in Macroecology and Biogeography"
subtitle: "Day 2: data.frames and GIS shapefiles in R; subsetting, joining & aggregating "
author: "Patrick Weigelt, Pierre Denelle, Amanda Taylor & Holger Kreft"
date: "April 27, 2022"
output: 
  html_document: 
    theme: flatly
    toc: yes
editor_options: 
 chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

&nbsp;
_[back to overview](index.html)_
&nbsp;
&nbsp;

**Preparation:** Please open the RStudio project we established last week and download all data for the class from  [Stud-IP](https://studip.uni-goettingen.de/dispatch.php/course/files/index/74501b33f332a6d238443bfbd502cc6c?cid=1746ae3fe971afbf90b74e3d64389586) (log in first) and unzip them into your project folder /data/. You can use `getwd()` to get your current working directory which is your project folder as long as you do not change it. Please install the following R-packages using `install.packages()` if not done yet:  

* sf
* raster
* dplyr

&nbsp;

If you want to visualize this tutorial in the viewer inside RStudio (to save space on your screen) run the following chunk of code:
```{r viewer, eval=FALSE}
install.packages("rstudioapi") # install an R-package required for this step
dir <- tempfile()
dir.create(dir)
download.file("https://gift.uni-goettingen.de/mcmmb/index.html", destfile = file.path(dir, "index.html"))
download.file("https://gift.uni-goettingen.de/mcmmb/Day2.html", destfile = file.path(dir, "Day2.html"))
htmlFile <- file.path(dir, "Day2.html")
rstudioapi::viewer(htmlFile)
```
Now you can conveniently copy code from the viewer into your script.

## 1. Loading spreadsheet data into R  

**Clean up**  
```{r}
rm(list = ls()) # remove all objects first; alternatively press CTRL+SHIFT+F10 to also unmap loaded packages
```
&nbsp;

**Load packages**  
```{r, include=FALSE}
library(dplyr)
```

```{r, eval=FALSE}
library(dplyr)
```
As a start, we use functions from base R to organize our data. However, we load the `dplyr` package, a toolkit for splitting, applying and combining data. For more advanced and faster functions related to data manipulation, have a look at the documentation of the packages `dplyr`, `reshape2`, `data.table` etc.  
&nbsp;

**Load data**  
We use a version of `read.table()` to import data into R as a data.frame. Have a look at its help page.  
```{r, eval=FALSE}
?read.table
```

```{r}
species <- read.csv("data/palms_species_per_gridcell.csv")
```
Here, we use data from [Kreft, H., Sommer, J.H. & Barthlott, W. (2006) The significance of geographic range size for spatial diversity patterns in Neotropical palms. _Ecography_, 29, 21-30](https://onlinelibrary.wiley.com/doi/full/10.1111/j.2005.0906-7590.04203.x).  
&nbsp;

**Inspection**
```{r}
dim(species)
colnames(species) # rownames(species) for rows
str(species)
```

```{r}
class(species$GENUS)
```

```{r}
unique(species$GENUS) # or levels(species$GENUS)
```


&nbsp;

**Adding a column to the data.frame**  
`paste()` allows us to combine both character strings of genus and epithet into a full species name.  
```{r}
species$species <- paste(species$GENUS, species$EPITHET, sep = " ")
```

&nbsp;

**Sorting**  
Click on the species object in the 'Environment' tab of RStudio or run `View(species)` to have a look at the table. Sorting in RStudio's viewer does not change the order of the actual data. To sort the data.frame, use `order()`.  

```{r, eval=FALSE}
order(species$grid_id)
```

Gives back the indices of grid_id in increasing order of the value of grid_id. They can be used to reorganize the data.frame accordingly

```{r}
species <- species[order(species$grid_id), ]
```
&nbsp;


## 2. Subsetting the data.frame using logical operators  
_`which()` and operators: `==`, `>=`, `>`, `<`, `<=`, `%in%`, `is.na()`, `is.finite()`, `!`, `&`, `|`_

**Select unique subset for certain columns**  
```{r}
species_unique <- unique(species[, c("spp_Id", "GENUS", "EPITHET", "species")])
# or: species_unique <- species[!duplicated(species$spp_Id),
# c("spp_Id", "GENUS", "EPITHET", "species")]

nrow(species_unique) # Number of species
length(unique(species$species))
```
&nbsp;

See how many species there are per genus
```{r}
spec_genera <- table(species_unique$GENUS) # contingency table of the counts of factor levels.
spec_genera

# Histogram
hist(spec_genera, main = "Histogram of species per genus",
     xlab = "Species per genus")
```
&nbsp;

**Choose entries of a given value**  
```{r}
which(species$GENUS == "Ammandra") # index of all entries of genus "Ammandra"
head(species[which(species$GENUS == "Ammandra"), ]) # first six rows

# What species belong to this genus?
unique(species$species[which(species$GENUS == "Ammandra")])
```
&nbsp;

**OR**  
```{r}
unique(species$species[which(species$GENUS == "Ammandra" | species$GENUS == "Washingtonia")]) # all species names in the genera "Ammandra" and "Washingtonia"
unique(species$species[which(species$GENUS %in% c("Ammandra", "Washingtonia"))])
```
&nbsp;

**AND**  
```{r}
head(species[which(species$GENUS=="Ammandra" & species$EPITHET == "decasperma"), ]) # all entries of Ammandra decasperma
head(species[which(species$species == "Ammandra decasperma"), ])
```
&nbsp;

## 3. data.frame exercise  
1) Make a subset of the data.frame that only includes entries for the 4 species _Geonoma pauciflora_, _Geonoma weberbaueri_, _Phytelephas tenuicaulis_ & _Mauritia flexuosa_.  

```{r, echo = FALSE, eval = FALSE}
focal_sp <- c("Geonoma pauciflora", "Geonoma weberbaueri",
              "Phytelephas tenuicaulis", "Mauritia flexuosa")

sub_sp <- species[which(species$species %in% focal_sp), ]
```

2) According to the [World Checklist of Vascular Plants](https://wcsp.science.kew.org/namedetail.do?name_id=89598), _Geonoma weberbaueri_ is a synonym of _Geonoma undata_. Rename it!

```{r, echo = FALSE, eval = FALSE}
unique(sub_sp$species)
sub_sp[which(sub_sp$species == "Geonoma weberbaueri"), "species"] <- "Geonoma undata"
unique(sub_sp$species)
```

3) In how many grid cells does each of these species occur?
&nbsp;

```{r, echo = FALSE, eval = FALSE}
table(sub_sp$species)
```

## 4. GIS shapefiles in R  

**Exercise: Have a look at the two shapefiles 30min_grid_select50%.shp and americas.shp in QGIS**  
&nbsp;

**Loading Geospatial R packages**
```{r, include=FALSE}
library(sf)
library(raster)
```

```{r, eval=FALSE}
install.packages("sf")
library(sf)
library(raster)
```

&nbsp;

**Load spatial poylgons into R**
```{r}
grid <- st_read("data/30min_grid_select50/30min_grid_select50%.shp")
americas <- st_read("data/americas/americas.shp")

# Plot
par(mfrow = c(1, 2))
plot(st_geometry(grid), border = "darkgrey")
plot(st_geometry(americas), add=TRUE)

plot(st_geometry(grid), border = "darkgrey", ylim = c(5, 25), xlim = c(-85, -65))
plot(st_geometry(americas), add = TRUE)
```

```{r}
class(grid)
dim(grid)
extent(grid)  
```
&nbsp;

**Projection**
```{r}
st_crs(grid)
```
not defined yet!
```{r}
st_crs(grid) <- "+proj=longlat +ellps=WGS84 +no_defs"
```

```{r}
st_crs(grid)
st_geometry(grid)
```
Check out <http://proj4.org/> or <http://epsg.io/> for spatial projections!  
&nbsp;

**Access the attribute table**  
```{r}
head(grid)
```
&nbsp;


# 5. Joining and aggregating information  

**Join species distributions to shapefile**

_Convert long table to species by grid-cell table_
```{r}
species <- species[, c("grid_id", "species")]
species_grid <- as.data.frame.matrix(table(species)) # Count species by gridcell combinations
species_grid[1:5, 1:5]

species_grid <- data.frame(grid_id = as.integer(row.names(species_grid)), species_grid) # Add column with grid_id
species_grid[1:5, 1:5]

species_grid[which(species_grid$grid_id %in% c(5328:5332)), 1:4]
```
&nbsp;

_Join palm occurrences and spatial shapefile_
```{r}
names(grid)[1] <- "grid_id" #rename ID column to correspond to species_grid$grid_id
grid <- left_join(grid, species_grid, by = "grid_id") # Perform a "SQL" left join
grid[is.na(grid)] <- 0
grid[which(grid$grid_id %in% c(5328:5332)), 1:5]
```
&nbsp;

**Make distribution maps for the four species**  
```{r}
plot(grid[c("Geonoma.pauciflora", "Geonoma.undata", "Phytelephas.tenuicaulis", "Mauritia.flexuosa")], border = NA, pal=c("grey","darkgreen"))
```
&nbsp;

```{r}
plot(st_geometry(grid), col=ifelse(grid$Geonoma.undata==1,"darkgreen","grey"), border=NA, main="Geonoma undata")
plot(st_geometry(americas), add=TRUE)
```
<br>

To go further, you can have a look at this nice tutorials: <http://edrub.in/ARE212/section12.html>,  <https://r-spatial.github.io/sf/articles/sf5.html>

