---
title: "Modern concepts and methods in Macroecology and Biogeography"
subtitle: "Day 2: data.frames and GIS shapefiles in R; subsetting, joining & aggregating "
author: "Dylan Craven & Patrick Weigelt"
date: "April 25, 2018"
output: 
  html_document: 
    theme: flatly
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Preperation:** Please open your RStudio project and download the data for today from [Stud-IP](https://www.studip.uni-goettingen.de/download/force_download/0/7cdab329853c1a57943e8406355d719e/Data_Day2.zip) (log in first) and unzip them into your project folder /data/. You can use `getwd()` to get your current working directory which is your project folder as long as you do not change it. Please install the following R-packages using `install.packages()` if not done yet:  

* rgdal  
* rgeos  
* maptools  
* raster  
* plyr  

&nbsp;

## 1. Loading spreadsheet data into R  

**Clean up**  
```{r}
rm(list=ls()) # remove all objects first; alternatively press CTRL+SHIFT+F10 to also unmap loaded packages
```
&nbsp;

**Load packages**  
```{r}
library(plyr) 
```
Here, we mostly use functions from base R to organize our data. However, we load the plyr package, a toolkit for splitting, applying and combining data. For more advanced and faster functions for data manipulation have a look at the packages dplyr, tidyr, reshape2, data.table etc.  
&nbsp;

**Load data**  
We use a version of `read.table()` to import data into R as a data.frame. Have a look at its help page  
```{r, eval=FALSE}
?read.table
```

```{r}
species <- read.csv("data/palms_species_per_gridcell.csv")
```
Here, we use data from [Kreft, H., Sommer, J.H. & Barthlott, W. (2006) The significance of geographic range size for spatial diversity patterns in Neotropical palms. _Ecography_, 29, 21-30](https://onlinelibrary.wiley.com/doi/full/10.1111/j.2005.0906-7590.04203.x).  
&nbsp;

**Inspection**
```{r}
dim(species)
colnames(species) # rownames(species) for rows
str(species)
```

```{r}
class(species$GENUS)
```
```{r}
unique(species$GENUS) # or levels(species$GENUS)
```
&nbsp;

**Adding a column to the data.frame**  
```{r}
species$species <- paste(species$GENUS,species$EPITHET, sep = " ")
```
`paste()` combines the characterstrings of genus and epithet into a full species name  
&nbsp;

**Sorting**  
Click on the species object in the 'Environment' tab of RStudio or run `View(species)` to have a look at the table. Sorting in RStudio's viewer does not change the order of the actual data. To sort the data.frame, use `order()`.  
```{r, eval=FALSE}
order(species$grid_id)
```
Gives back the indices of grid_id in increasing order of the value of grid_id. They can be used to reorganize the data.frame accordingly
```{r}
species <- species[order(species$grid_id),]
```
&nbsp;


## 2. Subsetting the data.frame using logical operators  
_`which()` and operators: `==`, `>=`, `>`, `<`, `<=`, `%in%`, `is.na()`, `is.finite()`, `!`, `&`, `|`_

**Select unique subset for certain columns**  
```{r}
species_unique <- unique(species[,c("spp_Id","GENUS","EPITHET","species")])
nrow(species_unique) # Number of species
```
&nbsp;

**Choose entries of a given value**  
```{r}
which(species$GENUS=="Ammandra") # index of all entries of genus "Ammandra"
head(species[which(species$GENUS=="Ammandra"),]) # first six rows
unique(species$species[which(species$GENUS=="Ammandra")])
```
&nbsp;

**OR**  
```{r}
unique(species$species[which(species$GENUS == "Ammandra" | species$GENUS == "Washingtonia")]) # all species names in the genera "Ammandra" and "Washingtonia"
unique(species$species[which(species$GENUS %in% c("Ammandra","Washingtonia"))])
```
&nbsp;

**AND**  
```{r}
head(species[which(species$GENUS=="Ammandra" & species$EPITHET =="decasperma"),]) # all entries of Ammandra decasperma
head(species[which(species$species=="Ammandra decasperma"),])
```
&nbsp;

## 3. data.frame exercise  
1) Make a subset the data.frame that only includes entries for the 4 species _Geonoma pauciflora_, _Geonoma weberbaueri_, _Phytelephas tenuicaulis_ & _Mauritia flexuosa_.  
2) According to [The Plant List](http://www.theplantlist.org/tpl1.1/record/kew-89598), _Geonoma weberbaueri_ is a synonym of _Geonoma undata_. Rename it!
3) In how many gridcells does each of the species occur?
&nbsp;


## 4. GIS shapefiles in R  

**Exercise: Have a look at the two shapefiles 30min_grid_select50%.shp and americas.shp in QGIS**  
&nbsp;

**Loading spatial packages**
```{r, include=FALSE}
library(rgdal)
library(rgeos)
library(maptools)
library(raster)
```
```{r, eval=FALSE}
library(rgdal)
library(rgeos)
library(maptools)
library(raster)
```

&nbsp;

**Load spatial poylgons into R**
```{r}
grid <- readOGR("data/30min_grid_select50%.shp", integer64="allow.loss")
americas <- readOGR("data/americas.shp")
par(mfrow=c(1,2))
plot(grid, border="darkgrey")
plot(americas, add=TRUE)
plot(grid, border="darkgrey", ylim=c(5,25), xlim=c(-85,-65))
plot(americas, add=TRUE)
```

```{r}
class(grid)
dim(grid)
extent(grid)  
```
&nbsp;

**Projection**
```{r}
grid@proj4string
```
not defined yet!
```{r}
grid@proj4string <- CRS("+proj=longlat +ellps=WGS84 +no_defs")
```
```{r}
grid@proj4string
```
Check <http://proj4.org/> for spatial projections!  
&nbsp;

**Access the attribute table**  
```{r}
head(grid@data)
```
&nbsp;


# 5. Joining and aggregating information  

**Join species distributions to shapefile**

_Convert long table to species by grid-cell table_
```{r}
species <- species[,c("grid_id","species")]
species_grid <- as.data.frame.matrix(table(species)) # Count species by gridcell combinations
species_grid <- data.frame(grid_id = as.integer(row.names(species_grid)), species_grid) # Add column with grid_id
names(grid@data)[1] <- "grid_id"
species_grid[1:5,1:4]
```
&nbsp;

_Join palm occurrences and spatial shapefile_
```{r}
grid@data <- join(grid@data, species_grid, by="grid_id", type="left") # Perform a "SQL" left join
grid@data[is.na(grid@data)] <- 0
grid@data[1:5,1:4]
```
&nbsp;

**Make distribution maps for the four species**  
```{r}
par(mfrow=c(2,2), mar=c(1,0,1,0))
plot(grid, col=ifelse(grid@data$Geonoma.pauciflora==1,"darkgreen","grey"), border=FALSE, main = "Geonoma pauciflora")
plot(grid, col=ifelse(grid@data$Geonoma.undata==1,"darkgreen","grey"), border=FALSE, main = "Geonoma undata")
plot(grid, col=ifelse(grid@data$Phytelephas.tenuicaulis==1,"darkgreen","grey"), border=FALSE, main = "Phytelephas tenuicaulis")
plot(grid, col=ifelse(grid@data$Mauritia.flexuosa==1,"darkgreen","grey"), border=FALSE, main = "Mauritia flexuosa")
```
&nbsp;

If you got interested have a look at this nice tutorial: <http://edrub.in/ARE212/section12.html>


<!-- ########################################################## -->
<!-- ## species numbers -->


<!-- # count species per gridcell -->
<!-- species_num <- data.frame(table(species$grid_id)) -->
<!-- head(species_num) -->
<!-- names(species_num) <- c("grid_id","spec_num") -->
<!-- head(species_num) -->

<!-- str(species_num) -->
<!-- species_num$grid_id <- as.numeric(as.character(species_num$grid_id)) -->
<!-- head(species_num) -->

<!-- str(species_num) -->
<!-- summary(species_num) -->


<!-- ## join species numbers to shapefile -->
<!-- names(grid@data)[1] <- "grid_id" -->
<!-- grid@data <- join(grid@data, species_num, by="grid_id", type="left") -->
<!-- head(grid@data) -->

<!-- ## Make a species richness map -->

<!-- #Create color scheme       -->
<!-- library(classInt) -->
<!-- library(colorRamps) -->

<!-- my.class.fr<-classIntervals(grid@data$spec_num,n=10,style="equal") # bin data into n quantiles -->
<!-- my.pal<-matlab.like(10) -->
<!-- my.col.fr<-findColours(my.class.fr,my.pal) # ramp colors based on classInts -->

<!-- plot(grid, col=my.col.fr, border=NA) -->




<!-- # histogram -->
<!-- hist(species_num$spec_num) -->

<!-- # number of grid cells with less or more than ten species -->
<!-- length(species_num$spec_num[which(species_num$spec_num > 10)]) -->
<!-- length(species_num$spec_num[which(!species_num$spec_num > 10)]) -->
<!-- length(species_num$spec_num[which(species_num$spec_num <= 10)]) -->



<!-- save(species_num, file="Day2/data/species_num.RData") -->



