---
title: 'Day 3 (alternate version): dplyr & ggplot2'
author: "Dylan Craven"
date: "May 2nd, 2018"
output: 
  html_document: 
    fig_height: 8
    fig_width: 8
    theme: flatly
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is an alternate version of the Day 3 practical that shows how to use other data manipulation and visualisation packages to achieve similar results.

### 1. Load packages  

```{r, include=FALSE}
library(dplyr)
library(tidyr)
library(reshape2)
library(ggplot2)
library(ggsci)
require(cowplot)
library(rgdal)
library(rgeos)
library(maptools)
library(raster)
library(BIEN)
library(ROCR)
```

**Load data and subset it**

```{r}
species <- read.csv("Data/palms_species_per_gridcell.csv")

species<- species %>%
          unite("species",c("GENUS","EPITHET"),sep=" ",remove=FALSE)

species$species<-as.character(species$species)

spp<-c("Phytelephas tenuicaulis","Mauritia flexuosa")

species<- filter(species,species %in% spp)
  
unique(species$species)

#species<- species%>% mutate(species = replace(species, species=="Geonoma weberbaueri", "Geonoma undata"))

```
&nbsp;

**Load GIS shapefile \& join it to species distribution data**  

```{r, echo=TRUE}
# read a polygon shapefile
grid <- readOGR("Data/30min_grid_select50%.shp", integer64="allow.loss")

grid@proj4string <- CRS("+proj=longlat +ellps=WGS84 +no_defs") # remember to include coordinate system

# make grid cell x species table
species$present<-1

species_grid<-dcast(species, grid_id~species,value.var="present",sum)

species_grid<-dplyr::select(species_grid, grid_id, Mauritia.flexuosa="Mauritia flexuosa",   
                     Phytelephas.tenuicaulis="Phytelephas tenuicaulis") # change column names

#
colnames(grid@data)[1] <- "grid_id"

grid@data <- left_join(grid@data, species_grid, by="grid_id")

grid@data<-grid@data %>% mutate_all(.funs = funs(ifelse(is.na(.), 0, .)))

grid@data[1:5,1:5]
```
&nbsp;

### 2. Download point occurences and load environmental layers and Americas shapefile


```{r, echo=TRUE}

spec_occ <- BIEN_occurrence_species(c("Phytelephas tenuicaulis","Mauritia flexuosa"), cultivated = FALSE, only.new.world = TRUE)

spec_occ <- filter(spec_occ, is.na(latitude)==FALSE & is.na(longitude)==FALSE)
```
&nbsp;


```{r, echo=TRUE}
elev <- raster("Data/mn30_grd")
bio1 <- raster("Data/CHELSA_bio10_01.tif") # temperature*10
bio12 <- raster("Data/CHELSA_bio10_12.tif")

```
&nbsp;


### 3. Pseudo-absences


**Make spatial points dataframe from grid**  

```{r, echo=TRUE}

grid_centroids <- gCentroid(grid, byid=TRUE)
grid_centroids <- SpatialPointsDataFrame(grid_centroids, data = data.frame(grid_centroids@coords, present = 0))
colnames(grid_centroids@data) <- c("longitude","latitude","present")

```
&nbsp;

**Make separate data frames with presences and pseudo-absences for each species**  


```{r, echo=TRUE}

# Mauritia flexuosa

mff<-filter(spec_occ, scrubbed_species_binomial=="Mauritia flexuosa")
mff$present<-1
mff<-dplyr::select(mff, longitude,latitude, present)
xy_1 <- mff[,c(1,2)] # coordinates

mau_fle_abs<- SpatialPointsDataFrame(coords =xy_1, data = mff,
                               proj4string = CRS("+proj=longlat +ellps=WGS84 +no_defs"))

# add pseudo-absences
set.seed(100)
mau_fle_abs <- rbind(grid_centroids[round(runif(nrow(mau_fle_abs), min = 1, max = nrow(grid_centroids))),],mau_fle_abs)


# Phytelephas tenuicaulis

ptt<-filter(spec_occ, scrubbed_species_binomial=="Phytelephas tenuicaulis")
ptt$present<-1
ptt<-dplyr::select(ptt, longitude,latitude, present)

xy_2 <- ptt[,c(1,2)] # coordinates

phy_ten_abs<- SpatialPointsDataFrame(coords =xy_2, data = ptt,
                               proj4string = CRS("+proj=longlat +ellps=WGS84 +no_defs"))

# add pseudo-absences
set.seed(100)
phy_ten_abs <- rbind(grid_centroids[round(runif(nrow(phy_ten_abs), min = 1, max = nrow(grid_centroids))),],phy_ten_abs)

```
&nbsp;

### 4. Fit species distribution models  


**Extract environmental data for presences/absences**

```{r, echo=TRUE}
# extract environmental data for presences and absences

phy_ten_abs@data$elev <- extract(elev,phy_ten_abs)
phy_ten_abs@data$bio1 <- extract(bio1,phy_ten_abs)
phy_ten_abs@data$bio12 <- extract(bio12,phy_ten_abs)

mau_fle_abs@data$elev <- extract(elev,mau_fle_abs)
mau_fle_abs@data$bio1 <- extract(bio1,mau_fle_abs)
mau_fle_abs@data$bio12 <- extract(bio12,mau_fle_abs)

```
&nbsp;



```{r, echo=TRUE}
logreg_phy_ten <- glm(present ~ elev + bio1 + bio12, data = phy_ten_abs@data, family = "binomial")
summary(logreg_phy_ten)

logreg_mau_fle <- glm(present ~ elev + bio1 + bio12, data = mau_fle_abs@data, family = "binomial")
summary(logreg_mau_fle)
```

### 5. Predict species distributions and compare to actual distributions

**Create new data for predictions**  


```{r, echo=TRUE}
new_extent <- extent(-97, -35, -30, 17) # crop the raster layers to smaller extent

elev <- crop(elev, new_extent) # crop raster layers to smaller extent
elev <- aggregate(elev, fact=4, fun=mean) # aggregate data

bio1 <- crop(bio1, new_extent)
bio1 <- aggregate(bio1, fact=4, fun=mean)

bio12 <- crop(bio12, new_extent)
bio12 <- aggregate(bio12, fact=4, fun=mean)

```
&nbsp;

**Predict presences of each species using species distribution models**  

We are using logistical regressions to fit species distribution models  


```{r, echo=TRUE}

new_data <- data.frame(elev = getValues(elev), bio1 = getValues(bio1), bio12 = getValues(bio12))

# predict presences
phy_ten_predict <- predict(logreg_phy_ten, newdata = new_data, type = "response")
mau_fle_predict <- predict(logreg_mau_fle, newdata = new_data, type = "response")


# create raster layers with SDM predictions
phy_ten_predict_rst <- raster(new_extent, nrows=nrow(elev), ncols=ncol(elev), vals = phy_ten_predict)
mau_fle_predict_rst <- raster(new_extent, nrows=nrow(elev), ncols=ncol(elev), vals = mau_fle_predict)

```
&nbsp;

**Make separate data frames for true species occurrences**    
```{r, echo=TRUE}

M_occ<-dplyr::filter(spec_occ, scrubbed_species_binomial=="Mauritia flexuosa")
M_occ$group<-1
P_occ<-dplyr::filter(spec_occ, scrubbed_species_binomial=="Phytelephas tenuicaulis")
P_occ$group<-1
```
&nbsp;

**Plot species ranges and occurrences using ggplot2 **  
```{r, echo=TRUE}

americas <- readOGR("Data/americas.shp") # add shape file showing the borders of the Neotropics  
americas@proj4string <- CRS("+proj=longlat +ellps=WGS84 +no_defs") 

americas_df <- ggplot2::fortify(americas) # so we can use it in ggplot2

data <- data.frame(id=rownames(grid@data), Mauritia.flexuosa=grid@data$Mauritia.flexuosa, Phytelephas.tenuicaulis=grid@data$Phytelephas.tenuicaulis,stringsAsFactors=F)
grid.df <- ggplot2::fortify(grid)
grid.df <- left_join(grid.df,data, by="id")

grid.df<-grid.df %>% mutate_all(.funs = funs(ifelse(is.na(.), 0, .)))

grid.df$Mauritia.flexuosa<-as.factor(grid.df$Mauritia.flexuosa)
grid.df$Phytelephas.tenuicaulis<-as.factor(grid.df$Phytelephas.tenuicaulis)


palm_Mauritia<-    ggplot(grid.df,aes(x=long,y=lat, group=group)) +  
                   geom_polygon(aes(fill=Mauritia.flexuosa))+ # polygon of species range
                   scale_fill_lancet()+
  
                  geom_polygon(data=americas_df, aes(x=long, y=lat,group=group), 
                  fill=NA, color="black", size=0.25) + # polygon of Americas
  
                  geom_point(data=M_occ,aes(x=longitude, y=latitude,group=group),shape=21, 
                             colour="white", fill="black",stroke=.2,size=2)+ #  occurrences from BIEN
  
                    coord_equal() + # same as coord_fixed(ratio = 1)
                   labs(x="Longitude",y="Latitude")+
                  ggtitle("Mauritia flexuosa")+
                  theme_bw() + theme(legend.position="none", 
                  plot.title = element_text(size=7,vjust=0.5),
                  axis.title=element_text(size=7, face="bold"),
                  axis.text.x=element_text(size=6), axis.text.y=element_text(size=6),
                  plot.margin =margin(t=0.02, r=0.1, b=0.01, l=0.1, unit="cm"))


palm_Phytelephas<-    ggplot(grid.df,aes(x=long,y=lat, group=group)) +  
                   geom_polygon(aes(fill=Phytelephas.tenuicaulis))+ # polygon of species range
                   scale_fill_lancet()+ 
  
                  geom_polygon(data=americas_df, aes(x=long, y=lat,group=group), 
                  fill=NA, color="black", size=0.25) + # polygon of Americas
  
                   geom_point(data=P_occ,aes(x=longitude, y=latitude,group=group),shape=21, 
                             colour="white", fill="black",stroke=.2,size=2)+ # occurrences from BIEN
  
                    coord_equal() + # same as coord_fixed(ratio = 1)
                   labs(x="Longitude",y="Latitude")+
                   ggtitle("Phytelephas tenuicaulis")+
                  theme_bw() + theme(legend.position="none", 
                  plot.title = element_text(size=7,vjust=0.5),
                  axis.title=element_text(size=7, face="bold"),
                  axis.text.x=element_text(size=6), axis.text.y=element_text(size=6),
                  plot.margin =margin(t=0.02, r=0.1, b=0.01, l=0.1, unit="cm"))


tog<-plot_grid(palm_Mauritia,palm_Phytelephas, ncol=2)

tog

```
&nbsp;

***Plot predicted ranges and occurrences**

```{r, echo=TRUE}


# make rasters using SDM predictions

phy_ten_predict_rst<- crop(x = phy_ten_predict_rst, y = new_extent,snap='in')
mau_fle_predict_rst<- crop(x = mau_fle_predict_rst, y = new_extent,snap='in')

# convert forest cover to Spatial data frame for ggplot2
phy_spdf <- as(phy_ten_predict_rst, "SpatialPixelsDataFrame")
phy_spdf <- as.data.frame(phy_spdf)
colnames(phy_spdf) <- c("value", "x", "y")

mau_spdf <- as(mau_fle_predict_rst, "SpatialPixelsDataFrame")
mau_spdf <- as.data.frame(mau_spdf)
colnames(mau_spdf) <- c("value", "x", "y")


# plot this

pred_Phytelephas<- ggplot(phy_spdf,aes(x=x,y=y, fill=value)) +

  geom_tile(alpha=0.8) +  # plot model predictions

  scale_fill_gradient(name="",low = "#00468BFF", high = "#ED0000FF")+
  
  geom_polygon(data=americas_df, aes(x=long, y=lat,group=group), 
                  fill=NA, color="black", size=0.25) + # polygon of Americas

 geom_point(data=P_occ,aes(x=longitude, y=latitude,group=group),shape=21, 
                             colour="white", fill="black",stroke=.2,size=2)+ # occurrences from BIEN
  
   coord_equal() + # same as coord_fixed(ratio = 1)
                   labs(x="Longitude",y="Latitude")+
                  theme_bw() + theme(legend.position="none", 
                  axis.title=element_text(size=7, face="bold"),
                  axis.text.x=element_text(size=6), axis.text.y=element_text(size=6),
                  plot.margin =margin(t=0.01, r=0.1, b=0.01, l=0.1, unit="cm"))

pred_Mauritia<-ggplot(mau_spdf,aes(x=x,y=y, fill=value)) +

  geom_tile(alpha=0.8) +  # plot model predictions

  scale_fill_gradient(name="",low = "#00468BFF", high = "#ED0000FF")+
  
  geom_polygon(data=americas_df, aes(x=long, y=lat,group=group), 
                  fill=NA, color="black", size=0.25) + # polygon of Americas

 geom_point(data=M_occ,aes(x=longitude, y=latitude,group=group),shape=21, 
                             colour="white", fill="black",stroke=.2,size=2)+ # occurrences from BIEN
  
   coord_equal() + # same as coord_fixed(ratio = 1)
                   labs(x="Longitude",y="Latitude")+
                  theme_bw() + theme(legend.position="none", 
                  axis.title=element_text(size=7, face="bold"),
                  axis.text.x=element_text(size=6), axis.text.y=element_text(size=6),
                  plot.margin =margin(t=0.01, r=0.1, b=0.01, l=0.1, unit="cm"))

tog2<-plot_grid(pred_Mauritia,pred_Phytelephas, ncol=2)

tog2
```
&nbsp;

**Combine figures**

```{r, echo=TRUE}

tog<-plot_grid(palm_Mauritia,palm_Phytelephas, ncol=2)
tog2<-plot_grid(pred_Mauritia,pred_Phytelephas, ncol=2)

togg3<-plot_grid(tog, tog2, ncol=1,align="hv")
togg3
```
&nbsp;

