---
title: "Modern concepts and methods in Macroecology and Biogeography"
subtitle: "Day 4: The latitudinal diversity gradient"
author: "Patrick Weigelt, Pierre Denelle & Holger Kreft"
date: "May 3rd, 2023"
output: 
  html_document: 
    theme: flatly
    toc: yes
editor_options: 
 chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c("top", "right"), color = "black")
```

&nbsp;
_[back to overview](index.html)_
&nbsp;
&nbsp;

**Preparation:** Open your RStudio project and download new data for today from [Stud-IP](https://studip.uni-goettingen.de/dispatch.php/course/files/index/74501b33f332a6d238443bfbd502cc6c?cid=1746ae3fe971afbf90b74e3d64389586) (log in first) and unzip them into your project folder '/data/. You can use `getwd()` to get your current working directory which is your project folder as long as you do not change it. Please install the following R-packages using `install.packages()` if not done yet:  

* `dplyr`  
* `classInt`
* `colorRamps`
* `performance`
* `sf`
* `ggplot2`
* `cowplot`

&nbsp;

If you want to visualize this tutorial in the viewer inside RStudio (to save space on your screen) run the following chunk of code:
```{r viewer, eval = FALSE}
install.packages("rstudioapi") # install an R-package required for this step
dir <- tempfile()
dir.create(dir)
download.file("https://gift.uni-goettingen.de/mcmmb/index.html",
              destfile = file.path(dir, "index.html"))
download.file("https://gift.uni-goettingen.de/mcmmb/Day4.html",
              destfile = file.path(dir, "Day4.html"))
htmlFile <- file.path(dir, "Day4.html")
rstudioapi::viewer(htmlFile)
```
Now you can conveniently copy code from the viewer into your script.

**Load R packages**  

```{r, include = FALSE}
library("dplyr")
library("classInt")
library("colorRamps")
library("performance")
library("sf")
library("ggplot2")
library("cowplot")
library("viridis")
```

```{r, eval = FALSE}
library("dplyr")
library("classInt")
library("colorRamps")
library("performance")
library("sf")
library("ggplot2")
library("cowplot")
library("viridis")
```

# 1. Loading palm data set and shapefiles

**Clean up workspace**  
```{r}
rm(list = ls()) # You might want to remove all loaded objects first
```
&nbsp;

**Load palm data set**  

For today's exercise, we will continue to use the Neotropical palm data set from [Kreft, H., Sommer, J.H. & Barthlott, W. (2006) The significance of geographic range size for spatial diversity patterns in Neotropical palms. _Ecography_, 29, 21-30](https://onlinelibrary.wiley.com/doi/full/10.1111/j.2005.0906-7590.04203.x).  
&nbsp;
```{r}
species <- read.csv("data/palms_species_per_gridcell.csv")
```
&nbsp;

**Data inspection**
```{r, echo = TRUE}
head(species)
str(species)
```
&nbsp;

Let's make a new column combining 'Genus' and 'Epithet'  
```{r, echo = TRUE}
species$species <- paste(species$GENUS,species$EPITHET, sep = " ") 
length(unique(species$species)) # number of species
n_distinct(species$species) # number of species with dplyr
```
&nbsp;

**Load GIS shapefiles**  
```{r, echo = TRUE}
# gridcells corresponding to palm distribution data and americas coastline
grid <- st_read("data/30min_grid_select50/30min_grid_select50%.shp")
americas <- st_read("data/americas/americas.shp")

# remember to define the Coordinate Reference System (CRS)
st_crs(grid) <- "+proj=longlat +ellps=WGS84 +no_defs"
st_crs(americas) <- "+proj=longlat +ellps=WGS84 +no_defs"

# have a look at the grid's attribute table
head(grid)

# change column name of first column ('ID')
colnames(grid)[colnames(grid) == "ID"] <- "grid_id"
head(grid)

# plot it
plot(st_geometry(grid), border = "darkgrey")
plot(st_geometry(americas), add = TRUE)

# with ggplot2
ggplot(grid) +
  geom_sf() +
  geom_sf(data = americas, fill = NA, color = "black") +
  ylim(-55, 40) + 
  theme_void()
```
&nbsp;

**Calculate latitude and longitude from the grid's polygons**
```{r, echo = TRUE, warning=FALSE}
# convert the polygons to points (mass centroids)
grid_centroids <- st_centroid(grid)

# extract the coordinates of the points
coordinates <- data.frame(st_coordinates(grid_centroids))
str(coordinates)

# add grid_id
coordinates <- data.frame(grid_id = grid_centroids$grid_id,
                          Long = coordinates$X,
                          Lat = coordinates$Y)

# join centroid coordinates to the grid shapefile
grid <- left_join(grid, coordinates, by = "grid_id")
head(grid)
```
&nbsp;

# 2. Data aggregation (species richness per gridcell)  

**Count species per gridcell**
```{r, echo = TRUE}
# ?table
species_num <- data.frame(table(species$grid_id))
head(species_num)
colnames(species_num) <- c("grid_id", "spec_num")
head(species_num)
str(species_num)

# change mode of grid_ID to numeric
species_num$grid_id <- as.numeric(as.character(species_num$grid_id))
str(species_num)

# Species richness distribution
summary(species_num$spec_num)

ggplot(species_num, aes(spec_num)) +
  geom_histogram(fill = "grey", color = "black") + # , bins = 30
  labs(x = "Species number", y = "Count") +
  theme_bw()
```
&nbsp;

**Join species richness per grid to the shapefile**  
```{r, echo = TRUE}
grid <- left_join(grid, species_num, by = "grid_id")
head(grid)
```
&nbsp;

**Make a species richness map**  
With a base R approach.
```{r}
# Defining a color palette
library("viridis")
cols <- viridis(n = 10)
colours <- colorRamp(cols)(grid$spec_num/max(grid$spec_num))
colours <- rgb(colours[,1], colours[,2], colours[,3], maxColorValue = 255)

# Plot
plot(st_geometry(grid),
     col = colours,
     # col = cols[grid$spec_num],
     border = NA,
     main = "Palm species richness")
plot(st_geometry(americas), add = TRUE)

legend_colours <- colorRamp(cols)(seq(0,1,length.out=8))
legend_colours <- rgb(legend_colours[,1], legend_colours[,2], 
                      legend_colours[,3], maxColorValue = 255)


legend("bottomleft", # position
       legend = round(seq(0,max(grid$spec_num),length.out=8)),
       title = "Species number",
       fill = legend_colours,
       cex = 0.7,
       bty = "n") # no box around it
```

With a ggplot2 approach.
```{r}
ggplot(grid) +
  geom_sf(color = NA, aes(fill = spec_num)) +
  geom_sf(data = americas, fill = NA, color = "black") +
  labs(title = "Palm species richness") +
  scale_fill_viridis_c("Species number") +
  ylim(-55, 40) +
  theme_void() + 
  theme(plot.title = element_text(hjust = 0.5,vjust = 10),
        plot.margin = margin(1,0,0,0, "cm"))
```

*Bonus: custom discrete color scale*
<details><summary>Bonus</summary>
Plot with intervals instead of a continuous color scale.
First with base R.
```{r, echo = TRUE}
# Splitting species richness in 10 intervals
my.class.fr <- classIntervals(grid$spec_num, n = 10, style = "equal",
                              dataPrecision = 0) # bin data into n quantiles
my.class.fr

# Palette of 10 colors
my.pal <- viridis(n = 10)
my.pal

# Create color scheme based on the intervals and the palette
my.col.fr <- findColours(my.class.fr, my.pal) # ramp colors based on classInts

# Plot
plot(st_geometry(grid), col = my.col.fr, border = NA,
     main = "Palm species richness")
plot(st_geometry(americas), add = TRUE)

legend("bottomleft", # position
       legend = names(attr(my.col.fr, "table")), 
       title = "Species number",
       fill = attr(my.col.fr, "palette"),
       cex = 0.7,
       bty = "n") # no box around it
```

Second with ggplot2.

```{r}
grid$spec_num_cat <- cut_interval(x = grid$spec_num, n = 10)

ggplot(grid) +
  geom_sf(color = NA, aes(fill = spec_num_cat)) +
  geom_sf(data = americas, fill = NA, color = "black") +
  labs(title = "Palm species richness") +
  scale_fill_viridis_d("Species number") +
  ylim(-55, 40) +
  theme_void() + 
  theme(plot.title = element_text(hjust = 0.5,vjust = 10),
        plot.margin = margin(1,0,0,0, "cm"))
```

</details>

# 3. Exercise  

1) How many of the grid cells are located within the Tropics (within +/- 23.433333)?  
```{r, echo = FALSE, eval = FALSE}
# length(grid@data$spec_num[which(grid$Lat > -23.433333 &
# grid$Lat < 23.433333)])
length(which(abs(grid$Lat) < 23.433333))
nrow(grid[which(abs(grid$Lat) < 23.433333), ])
```
&nbsp;

2) Compare mean species richness per gridcell within and outside the Tropics  
```{r, echo = FALSE, eval = FALSE}
data.frame(
  Tropics = mean(grid$spec_num[which(abs(grid$Lat) < 23.433333)]),
  nonTropics = mean(grid$spec_num[which(abs(grid$Lat) >= 23.433333)]))

# alternatively
grid$climate <- ifelse(grid$Lat > -23.433333 & grid$Lat < 23.433333, 
                       "tropical", "non-tropical")
aggregate(grid$spec_num, list(grid$climate), FUN = mean)

```
&nbsp;

3) How many grid cells have more than 20 species and are located South of 17°S? Can you plot them?
```{r, echo = FALSE, eval = FALSE}
length(grid$spec_num[which(grid$spec_num > 20 & grid$Lat < -17)])

plot(st_geometry(grid), border= "darkgrey")
plot(st_geometry(grid[which(grid$spec_num > 20 & grid$Lat < -17), ]),
     col = "red", border = NA, add = TRUE)
plot(st_geometry(americas), add = TRUE)

# With ggplot2
ggplot(grid) +
  geom_sf() +
  geom_sf(data = grid[which(grid$spec_num > 20 & grid$Lat < -17), ],
          color = "firebrick3") +
  geom_sf(data = americas, color = "black", fill = NA) +
  ylim(-55, 40) +
  theme_void()
```

# 4. Data aggregation (species richness per latitudinal belt)  

```{r, echo = TRUE}
# Merging coordinates to species_num data.frame
species_num <- inner_join(species_num, coordinates, by = "grid_id")
range(species_num$Lat)
```
&nbsp;

**Establish latitudinal belts**
```{r, echo = TRUE}
belts <- seq(round(min(species_num$Lat),2), round(max(species_num$Lat),2), by = 0.5)
belts
length(belts)
```
&nbsp;

**Species richness of one particular latitudinal belt**  
Select gridcells based on latitude
```{r, echo = TRUE}
species_num[which(species_num$Lat > -25.5 & species_num$Lat < -25),
            "spec_num"]
```
&nbsp;

Select gridcells based on latitude taking from belts object
```{r, echo = TRUE}
belts[20]
species_num[which(species_num$Lat > belts[20] - 0.25 &
                    species_num$Lat < belts[20] + 0.25), "spec_num"]

mean(species_num[which(species_num$Lat > belts[20] - 0.25 &
                         species_num$Lat < belts[20] + 0.25), "spec_num"])
```
&nbsp;

**Species richness across all latitudinal belts ('for loop approach')**
```{r, echo = TRUE}
# for-loop example
for (i in 1:5){
  print(i)
  Sys.sleep(1)
}

# create an empty vector for the mean species counts
species_per_belt_loop <- NA

# loop trough the latitudinal belts and calculate the mean species richness
for (i in 1:length(belts)){
  belt_subset <- species_num[which(species_num$Lat > belts[i] - 0.25 &
                                     species_num$Lat < belts[i] + 0.25),
                             "spec_num"]
  species_per_belt_loop[i] <- mean(belt_subset)
}
```
&nbsp;

**Species richness across all latitudinal belts ('apply a function approach')**
```{r, echo = TRUE}
# function example
myFunction <- function(x){
  if (x > 5) message("Hello")
}
myFunction(5)
myFunction(6)

# define the function
species_per_belt_function <- function(x){
  mean(species_num[which(species_num$Lat > x - 0.25 &
                           species_num$Lat < x + 0.25), "spec_num"])
}

# apply the function to one latitudinal belt
species_per_belt_function(-15.75)

# apply the function to all belts
species_per_belt_apply <- sapply(belts, function(x)
  mean(species_num$spec_num[which(species_num$Lat > x - 0.25 &
                                    species_num$Lat < x + 0.25)]))

species_per_belt_apply <- sapply(belts, species_per_belt_function)
```
&nbsp;

**Species richness across all latitudinal belts ('dplyr approach')**
```{r, echo = TRUE}
# create belt column
species_num$belt <- round(species_num$Lat*100/25)*25/100 # round at a quarter
species_num <- group_by(species_num, belt)
species_per_belt_dplyr <- summarize(species_num, richness = mean(spec_num))
species_per_belt_dplyr <- species_per_belt_dplyr$richness
species_num <- ungroup(species_num)
```
&nbsp;

**Compare loop, apply and dplyr approach**
```{r, echo = TRUE}
# compare loop and apply approach
all(species_per_belt_apply == species_per_belt_loop)
all(species_per_belt_apply == species_per_belt_dplyr)

species_per_belt <- species_per_belt_apply
```
&nbsp;

**Save the results**
```{r, echo = TRUE, eval = FALSE}
save(species_num, file = "data/species_num.RData")
save(belts, species_per_belt, file = "data/species_num_belts.RData")
write.csv(species_num, "data/species_num.csv", row.names = FALSE)
```
&nbsp;

# 5. The latitudinal gradient model (LDG)
In case you had problems creating the `species_num` and `species_per_belt` data.frames you can now load our provided *.Rdata files
```{r, echo = TRUE}
load("data/species_num.RData")
load("data/species_num_belts.RData")
```
&nbsp;

**Gridcell level**  
```{r, echo = TRUE}
par(mfrow = c(1, 2))
plot(species_num$Lat, species_num$spec_num,
     xlab = "Latitude", ylab = "Species number",
     pch = 16, col = "#00000070") 
plot(abs(species_num$Lat), species_num$spec_num,
     xlab = "Absolute latitude", ylab = "Species number",
     pch = 16, col = "#00000070")  
# for color codes, see here: https://htmlcolorcodes.com/

# With ggplot2
plot_grid(
  ggplot(species_num, aes(Lat, spec_num)) +
    geom_point(color = "black", alpha = 0.3) +
    labs(x = "Latitude", y = "Species number") +
    theme_bw(),
  ggplot(species_num, aes(abs(Lat), spec_num)) +
    geom_point(color = "black", alpha = 0.3) +
    labs(x = "Absolute latitude", y = "Species number") +
    theme_bw()
)
```  
&nbsp;  
What could be the reason for the triangular relationship?  
&nbsp;

**Linear model of species richness ~ latitude**
```{r, echo = TRUE}
# run a linear model for palm species richness in dependence on latitude
model1 <- lm(species_num$spec_num ~ abs(species_num$Lat))
summary(model1)

par(mfrow = c(1, 1))
plot(abs(species_num$Lat), species_num$spec_num,
     xlab = "Absolute latitude", ylab = "Species number",
     pch = 16, col = "#00000050")
abline(model1, col = "red", lwd = 2)

# With ggplot2
ggplot(species_num, aes(abs(Lat), spec_num)) +
  geom_point(color = "black", alpha = 0.3) +
  stat_smooth(method = "lm", color = "red") +
  labs(x = "Absolute latitude", y = "Species number") +
  theme_bw()
```
&nbsp;

**Latitudinal belts**
```{r, echo = TRUE}
model2 <- lm(species_per_belt ~ abs(belts))
summary(model2)

model3 <- lm(species_per_belt ~ abs(belts) + I(abs(belts)^2))
summary(model3)

# Plot
par(mfrow = c(1, 2))
plot(belts, species_per_belt,
     xlab = "Latitude", ylab = "Species number", pch = 16)
plot(abs(belts), species_per_belt,
     xlab = "Absolute latitude", ylab = "Species number",pch = 16)
abline(model2, col = "red", lwd = 2)
points(abs(belts), predict(model3), type = "l", col = "blue", lwd = 2)

# With ggplot2
belt_sp <- data.frame(belts = belts,
                      spec_num = species_per_belt)

plot_grid(
  ggplot(belt_sp, aes(belts, spec_num)) +
    geom_point(color = "black") +
    labs(x = "Latitude", y = "Species number") +
    theme_bw(),
  ggplot(belt_sp, aes(abs(belts), spec_num)) +
    geom_point(color = "black") +
    stat_smooth(method = "lm", formula = y ~ abs(x),
                color = "red") +
    stat_smooth(method = "lm", formula = y ~ abs(x) + I(abs(x)^2),
                color = "blue") +
    labs(x = "Absolute latitude", y = "Species number") +
    theme_bw()
)
```

&nbsp;

Let's also assess model performance using the performance package to calculate commonly used indices, such as R^2^, AIC, RMSE, etc.  

```{r, echo = TRUE}
AIC(model2, model3)
compare_performance(model3, model2)
```
