---
title: "Modern concepts and methods in Macroecology and Biogeography"
subtitle: "Day 4: The latitudinal diversity gradient"
author: "Dylan Craven & Patrick Weigelt"
date: "May 9th, 2018"
output: 
  html_document: 
    theme: flatly
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Preparation:** Please open your RStudio project and download the data for today from [Stud-IP](https://www.studip.uni-goettingen.de/dispatch.php/file/details/b64b68db7a05f02fa854a63a01064f67?cid=95a0e5be352a34697791cddd2f4ddbf4) (log in first) and unzip them into your project folder '/data/. You can use `getwd()` to get your current working directory which is your project folder as long as you do not change it. Please install the following R-packages using `install.packages()` if not done yet:  

* rgdal  
* rgeos  
* maptools  
* plyr  
* classInt
* colorRamps

&nbsp;

### 1. Loading palm data set and shapefiles

**Clean up workspace**  
```{r}
rm(list=ls()) # You might want to remove all loaded objects first
```
&nbsp;

**Load R packages & palm data set**  


```{r, include=FALSE}
library(plyr)
library(rgdal)
library(rgeos)
library(maptools)
library(classInt)
library(colorRamps)
```

```{r, eval=FALSE}
library(plyr)
library(rgdal)
library(rgeos)
library(maptools)
library(classInt)
library(colorRamps)
```
&nbsp;

```{r}
species <- read.csv("data/palms_species_per_gridcell.csv")
```
For today's exercise, we will continue to use the Neotropical palm data set from [Kreft, H., Sommer, J.H. & Barthlott, W. (2006) The significance of geographic range size for spatial diversity patterns in Neotropical palms. _Ecography_, 29, 21-30](https://onlinelibrary.wiley.com/doi/full/10.1111/j.2005.0906-7590.04203.x).  
&nbsp;


**Data inspection**
```{r, echo=TRUE}
head(species) 
str(species)
```
&nbsp;

Let's make a new column combining 'Genus' and 'Epithet'  
```{r, echo=TRUE}
species$species <- paste(species$GENUS,species$EPITHET, sep = " ") 
length(unique(species$species)) # number of species
```
&nbsp;



**Load GIS shapefiles**  
```{r, echo=TRUE}
# gridcells corresponding to palm distribution data and americas coastline
grid <- readOGR("Data/30min_grid_select50%.shp", integer64="allow.loss")
americas <- readOGR("Data/americas.shp")

# remember to include coordinate system
grid@proj4string <- CRS("+proj=longlat +ellps=WGS84 +no_defs")
americas@proj4string <- CRS("+proj=longlat +ellps=WGS84 +no_defs")

# have a look at the grid's attribute table
head(grid@data) 

# change clomumn name of first column ('ID')
names(grid@data)[1] <- "grid_id"
head(grid@data) 

# plot it
plot(grid, border= "darkgrey")
plot(americas, add=TRUE)
```
&nbsp;


**Calculate latitude and longitude from the grid's polygons**
```{r, echo=TRUE}
# convert the polygons to points (mass centroids)
grid_centroids <- gCentroid(grid, byid=TRUE)

# extract the coordinates of the points
coordinates <- data.frame(grid_centroids@coords)
head(coordinates)

# add grid_id to coordinates data.frame
coordinates <- data.frame(grid_id = grid@data$grid_id, Long = coordinates$x, Lat = coordinates$y)

# join coordinates to grid shapefile
grid@data <- join(grid@data, coordinates, by="grid_id", type="left")
```
&nbsp;


# 2. Data aggregation (species richness per gridcell)  

**Count species per gridcell**
```{r, echo=TRUE}
?table
species_num <- data.frame(table(species$grid_id))
head(species_num)
names(species_num) <- c("grid_id","spec_num")
head(species_num)
str(species_num)

# change mode of grid_ID to numeric
species_num$grid_id <- as.numeric(as.character(species_num$grid_id))
str(species_num)
summary(species_num)
```
&nbsp;

**Join species numbers to shapefile**  
```{r, echo=TRUE}
grid@data <- join(grid@data, species_num, by="grid_id", type="left")
head(grid@data)
```
&nbsp;

**Make a species richness map** 
```{r, echo=TRUE}
#Create color scheme
my.class.fr<-classIntervals(grid@data$spec_num, n=10, style="equal", dataPrecision=0) # bin data into n quantiles
my.class.fr
my.pal<-matlab.like(10)
my.col.fr<-findColours(my.class.fr,my.pal) # ramp colors based on classInts

plot(grid, col=my.col.fr, border=NA, main="Palm species richness")
plot(americas, add=TRUE)

legend("bottomleft", # position
  legend = names(attr(my.col.fr, "table")), 
  title = "Species number",
  fill = attr(my.col.fr, "palette"),
  cex = 0.7,
  bty = "n") # no box around it
```


# 3. Exercise  

1) How many of the grid cells are located within the Tropics (within +/- 23.433333?)  
```{r, echo=FALSE}
#length(grid@data$spec_num[which(grid@data$Lat > -23.433333 & grid@data$Lat < 23.433333)])
length(which(abs(grid@data$Lat) < 23.433333))
```
&nbsp;

2) Compare mean species richness per gridcell within and outside the Tropics  
```{r, echo=FALSE}
data.frame(Tropics=mean(grid@data$spec_num[which(abs(grid@data$Lat) < 23.433333)]),
           nonTropics=mean(grid@data$spec_num[which(abs(grid@data$Lat) >= 23.433333)]))
```
&nbsp;

3) How many gridcells have more than 20 species and are located South of 17°S? Can you plot them?
```{r, echo=FALSE}
length(grid@data$spec_num[which(grid@data$spec_num > 20 &  grid@data$Lat < -17)])

plot(grid, border= "darkgrey")
plot(grid[which(grid@data$spec_num > 20 & grid@data$Lat < -17),], col="red",
     border=NA, add=TRUE)
plot(americas, add=TRUE)

```


# 4. Data aggregation (species richness per latitudinal belt)  

```{r, echo=TRUE}
species_num <- join(species_num, coordinates, by="grid_id", type="inner")
min(species_num$Lat);max(species_num$Lat)
```
&nbsp;

**Establish latitudinal belts**
```{r, echo=TRUE}
belts <- seq(-34.75, 36.25, by = 0.5)
belts
```
&nbsp;


**Species richness of one particular latitudinal belt**  
Select gridcells based on latitude
```{r, echo=TRUE}
species_num$spec_num[which(species_num$Lat>-25.5 & species_num$Lat< -25)]
```
&nbsp;


Select gridcells based on latitude taking from belts object
```{r, echo=TRUE}
species_num$spec_num[which(species_num$Lat>belts[20]-0.25 &
                                            species_num$Lat<belts[20]+0.25)]

mean(species_num$spec_num[which(species_num$Lat>belts[20]-0.25 &
                                  species_num$Lat<belts[20]+0.25)])

```
&nbsp;


**Species richness across all latitudinal belts ('for loop approach')**
```{r, echo=TRUE}
#for loop example
for (i in 1:10){
  print(i)
}

# create an empty vector for the mean species counts
species_per_belt_loop <- NA

# loop trough the latitudinal belts and calculate the mean species richness
for (i in 1:length(belts)){
  belt_subset <- species_num$spec_num[which(species_num$Lat>belts[i]-0.25 & species_num$Lat<belts[i]+0.25)]
  species_per_belt_loop[i] <- mean(belt_subset)
}

```
&nbsp;


**Species richness across all latitudinal belts ('apply a function approach')**
```{r, echo=TRUE}

# function example
myFunction <- function(x){
  if (x > 5) message("Hello")
}

myFunction(6)

# define the function
species_per_belt_function <- function(x) {
  mean(species_num$spec_num[which(species_num$Lat>x-0.25 & species_num$Lat<x+0.25)])
}

# apply the function to one latitudinal belt
species_per_belt_function(-15.75)

# apply the function to all belts
species_per_belt_apply <- sapply(belts, function(x)
  mean(species_num$spec_num[which(species_num$Lat>x-0.25 & species_num$Lat<x+0.25)]))

species_per_belt_apply <- sapply(belts, species_per_belt_function)
```
&nbsp;


**Compare loop and apply approach**
```{r, echo=TRUE}
# compare loop and apply approach
all(species_per_belt_apply == species_per_belt_loop)
species_per_belt <- species_per_belt_apply
```
&nbsp;


**Save the results**
```{r, echo=TRUE}
save(species_num, file="data/species_num.RData")
save(belts, species_per_belt, file="data/species_num_belts.RData")
write.csv(species_num, "data/species_num.csv", row.names = FALSE)
```
&nbsp;


# 5. The latitudinal gradient model  

In case you had problems creating the `species_num` and `species_per_belt` data.frames you can now load our provided *.Rdata files
```{r, echo=TRUE}
load("data/species_num.RData")
load("data/species_num_belts.RData")
```
&nbsp;


**Gridcell level**  
```{r, echo=TRUE}
par(mfrow=c(1,2))
plot(species_num$Lat,species_num$spec_num, xlab="Latitude", ylab="Species number") 
plot(abs(species_num$Lat),species_num$spec_num, xlab="Absolute latitude", ylab="Species number")  
```  
&nbsp;  
What could be the reason for the triangular relationship?  
&nbsp;

**Linear model of species richness ~ latitude**
```{r, echo=TRUE}
# run a linear model for palm species richness in dependence on latitude
model1 <- lm(species_num$spec_num ~ abs(species_num$Lat))
summary(model1)

plot(abs(species_num$Lat),species_num$spec_num, xlab="Absolute latitude", ylab="Species number")
abline(model1, col="red", lwd=2)
```
&nbsp;

**Latitudinal belts**
```{r, echo=TRUE}
model2 <- lm(species_per_belt_loop ~ abs(belts))
summary(model2)

model3 <- lm(species_per_belt_loop ~ abs(belts) + I(abs(belts)^2))
summary(model3)

par(mfrow=c(1,2))
plot(belts,species_per_belt_loop, xlab="Latitude", ylab="Species number")
plot(abs(belts),species_per_belt_loop, xlab="Absolute latitude", ylab="Species number")
abline(model2, col="red", lwd=2)
points(abs(belts), predict(model3), type="l", col="blue", lwd=2)
AIC(model2,model3)
```



