---
title: "Modern concepts and methods in Macroecology and Biogeography"
subtitle: "Day 4: The latitudinal diversity gradient"
author: "Patrick Weigelt, Pierre Denelle, Amanda Taylor & Holger Kreft"
date: "May 5th, 2021"
output: 
  html_document: 
    theme: flatly
    toc: yes
editor_options: 
 chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

&nbsp;
_[back to overview](index.html)_
&nbsp;
&nbsp;


**Preparation:** Open your RStudio project and download new data for today from [Stud-IP](https://studip.uni-goettingen.de/dispatch.php/course/files/index/74501b33f332a6d238443bfbd502cc6c?cid=1746ae3fe971afbf90b74e3d64389586) (log in first) and unzip them into your project folder '/data/. You can use `getwd()` to get your current working directory which is your project folder as long as you do not change it. Please install the following R-packages using `install.packages()` if not done yet:  

* `rgdal`  
* `rgeos`  
* `dplyr`  
* `classInt`
* `colorRamps`
* `performance`

&nbsp;

If you want to visualize this tutorial in the viewer inside RStudio (to save space on your screen) run the following chunk of code:
```{r viewer, eval = FALSE}
install.packages("rstudioapi") # install an R-package required for this step
dir <- tempfile()
dir.create(dir)
download.file("https://gift.uni-goettingen.de/mcmmb/index.html",
              destfile = file.path(dir, "index.html"))
download.file("https://gift.uni-goettingen.de/mcmmb/Day4.html",
              destfile = file.path(dir, "Day4.html"))
htmlFile <- file.path(dir, "Day4.html")
rstudioapi::viewer(htmlFile)
```
Now you can conveniently copy code from the viewer into your script.

### 1. Loading palm data set and shapefiles

**Clean up workspace**  
```{r}
rm(list = ls()) # You might want to remove all loaded objects first
```
&nbsp;

**Load R packages & palm data set**  


```{r, include = FALSE}
library(dplyr)
library(rgdal)
library(rgeos)
library(classInt)
library(colorRamps)
library(performance)
```

```{r, eval = FALSE}
library(dplyr)
library(rgdal)
library(rgeos)
library(classInt)
library(colorRamps)
library(performance)
```
&nbsp;

For today's exercise, we will continue to use the Neotropical palm data set from [Kreft, H., Sommer, J.H. & Barthlott, W. (2006) The significance of geographic range size for spatial diversity patterns in Neotropical palms. _Ecography_, 29, 21-30](https://onlinelibrary.wiley.com/doi/full/10.1111/j.2005.0906-7590.04203.x).  
&nbsp;
```{r}
species <- read.csv("data/palms_species_per_gridcell.csv")
```
&nbsp;


**Data inspection**
```{r, echo = TRUE}
head(species)
str(species)
```
&nbsp;

Let's make a new column combining 'Genus' and 'Epithet'  
```{r, echo = TRUE}
species$species <- paste(species$GENUS,species$EPITHET, sep = " ") 
length(unique(species$species)) # number of species
```
&nbsp;


**Load GIS shapefiles**  
```{r, echo = TRUE}
# gridcells corresponding to palm distribution data and americas coastline
grid <- readOGR("data/30min_grid_select50/30min_grid_select50%.shp",
                integer64 = "allow.loss")
americas <- readOGR("data/americas/americas.shp")

# remember to include coordinate system
grid@proj4string <- CRS("+proj=longlat +ellps=WGS84 +no_defs")
americas@proj4string <- CRS("+proj=longlat +ellps=WGS84 +no_defs")

# have a look at the grid's attribute table
head(grid@data) 

# change clomumn name of first column ('ID')
names(grid@data)[1] <- "grid_id"
head(grid@data) 

# plot it
plot(grid, border= "darkgrey")
plot(americas, add = TRUE)
```
&nbsp;

**Calculate latitude and longitude from the grid's polygons**
```{r, echo = TRUE}
# convert the polygons to points (mass centroids)
grid_centroids <- gCentroid(grid, byid = TRUE)

# extract the coordinates of the points
coordinates <- data.frame(grid_centroids@coords)
head(coordinates)

# add grid_id to coordinates data.frame
coordinates <- data.frame(grid_id = grid@data$grid_id,
                          Long = coordinates$x, Lat = coordinates$y)

# join centroid coordinates to the grid shapefile
grid@data <- left_join(grid@data, coordinates, by = "grid_id")
head(grid@data)
```
&nbsp;

### 2. Data aggregation (species richness per gridcell)  

**Count species per gridcell**
```{r, echo = TRUE}
?table
species_num <- data.frame(table(species$grid_id))
head(species_num)
colnames(species_num) <- c("grid_id", "spec_num")
head(species_num)
str(species_num)

# change mode of grid_ID to numeric
species_num$grid_id <- as.numeric(as.character(species_num$grid_id))
str(species_num)
summary(species_num)
```
&nbsp;

**Join species richness per grid to the shapefile**  
```{r, echo = TRUE}
grid@data <- left_join(grid@data, species_num, by = "grid_id")
head(grid@data)
```
&nbsp;

**Make a species richness map** 
```{r, echo = TRUE}
# Splitting species richness in 10 intervals
my.class.fr <- classIntervals(grid@data$spec_num, n = 10, style = "equal",
                              dataPrecision = 0) # bin data into n quantiles
my.class.fr

# Palette of 10 colors
my.pal <- matlab.like(10)
my.pal

# Create color scheme based on the intervals and the palette
my.col.fr <- findColours(my.class.fr, my.pal) # ramp colors based on classInts

# Plot
plot(grid, col = my.col.fr, border = NA, main = "Palm species richness")
plot(americas, add = TRUE)

legend("bottomleft", # position
       legend = names(attr(my.col.fr, "table")), 
       title = "Species number",
       fill = attr(my.col.fr, "palette"),
       cex = 0.7,
       bty = "n") # no box around it
```


### 3. Exercise  

1) How many of the grid cells are located within the Tropics (within +/- 23.433333)?  
```{r, echo = FALSE, eval = FALSE}
# length(grid@data$spec_num[which(grid@data$Lat > -23.433333 &
# grid@data$Lat < 23.433333)])
length(which(abs(grid@data$Lat) < 23.433333))
```
&nbsp;

2) Compare mean species richness per gridcell within and outside the Tropics  
```{r, echo = FALSE, eval = FALSE}
data.frame(Tropics = mean(grid@data$spec_num[which(abs(grid@data$Lat) < 23.433333)]),
           nonTropics = mean(grid@data$spec_num[which(abs(grid@data$Lat) >= 23.433333)]))
```
&nbsp;

3) How many gridcells have more than 20 species and are located South of 17°S? Can you plot them?
```{r, echo = FALSE, eval = FALSE}
length(grid@data$spec_num[which(grid@data$spec_num > 20 &
                                  grid@data$Lat < -17)])

plot(grid, border= "darkgrey")
plot(grid[which(grid@data$spec_num > 20 & grid@data$Lat < -17), ],
     col="red", border = NA, add = TRUE)
plot(americas, add = TRUE)
```

### 4. Data aggregation (species richness per latitudinal belt)  

```{r, echo = TRUE}
# Merging coordinates to species_num data.frame
species_num <- inner_join(species_num, coordinates, by = "grid_id")
range(species_num$Lat)
```
&nbsp;

**Establish latitudinal belts**
```{r, echo = TRUE}
belts <- seq(min(species_num$Lat), max(species_num$Lat), by = 0.5)
belts
```
&nbsp;

**Species richness of one particular latitudinal belt**  
Select gridcells based on latitude
```{r, echo = TRUE}
species_num[which(species_num$Lat > -25.5 & species_num$Lat < -25),
            "spec_num"]
```
&nbsp;

Select gridcells based on latitude taking from belts object
```{r, echo = TRUE}
belts[20]
species_num[which(species_num$Lat > belts[20] - 0.25 &
                    species_num$Lat < belts[20] + 0.25), "spec_num"]

mean(species_num[which(species_num$Lat > belts[20] - 0.25 &
                         species_num$Lat < belts[20] + 0.25), "spec_num"])
```
&nbsp;

**Species richness across all latitudinal belts ('for loop approach')**
```{r, echo = TRUE}
#for loop example
for (i in 1:5){
  print(i)
  Sys.sleep(1)
}

# create an empty vector for the mean species counts
species_per_belt_loop <- NA

# loop trough the latitudinal belts and calculate the mean species richness
for (i in 1:length(belts)){
  belt_subset <- species_num[which(species_num$Lat > belts[i] - 0.25 &
                                     species_num$Lat<belts[i] + 0.25),
                             "spec_num"]
  species_per_belt_loop[i] <- mean(belt_subset)
}

```
&nbsp;

**Species richness across all latitudinal belts ('apply a function approach')**
```{r, echo = TRUE}
# function example
myFunction <- function(x){
  if (x > 5) message("Hello")
}
myFunction(5)
myFunction(6)

# define the function
species_per_belt_function <- function(x){
  mean(species_num[which(species_num$Lat > x - 0.25 &
                           species_num$Lat < x + 0.25), "spec_num"])
}

# apply the function to one latitudinal belt
species_per_belt_function(-15.75)

# apply the function to all belts
species_per_belt_apply <- sapply(belts, function(x)
  mean(species_num$spec_num[which(species_num$Lat > x - 0.25 &
                                    species_num$Lat < x + 0.25)]))

species_per_belt_apply <- sapply(belts, species_per_belt_function)
```
&nbsp;


**Compare loop and apply approach**
```{r, echo = TRUE}
# compare loop and apply approach
all(species_per_belt_apply == species_per_belt_loop)
species_per_belt <- species_per_belt_apply
```
&nbsp;

**Save the results**
```{r, echo = TRUE, eval = FALSE}
save(species_num, file = "data/species_num.RData")
save(belts, species_per_belt, file = "data/species_num_belts.RData")
write.csv(species_num, "data/species_num.csv", row.names = FALSE)
```
&nbsp;

### 5. The latitudinal gradient model  
In case you had problems creating the `species_num` and `species_per_belt` data.frames you can now load our provided *.Rdata files
```{r, echo = TRUE}
load("data/species_num.RData")
load("data/species_num_belts.RData")
```
&nbsp;

**Gridcell level**  
```{r, echo = TRUE}
par(mfrow = c(1, 2))
plot(species_num$Lat, species_num$spec_num,
     xlab = "Latitude", ylab = "Species number",
     pch = 16, col = "#00000070") 
plot(abs(species_num$Lat), species_num$spec_num,
     xlab = "Absolute latitude", ylab = "Species number",
     pch = 16, col = "#00000070")  
# for color codes, see here: https://htmlcolorcodes.com/
```  
&nbsp;  
What could be the reason for the triangular relationship?  
&nbsp;

**Linear model of species richness ~ latitude**
```{r, echo = TRUE}
# run a linear model for palm species richness in dependence on latitude
model1 <- lm(species_num$spec_num ~ abs(species_num$Lat))
summary(model1)

par(mfrow = c(1, 1))
plot(abs(species_num$Lat), species_num$spec_num,
     xlab = "Absolute latitude", ylab = "Species number",
     pch = 16, col = "#00000050")
abline(model1, col = "red", lwd = 2)
```
&nbsp;

**Latitudinal belts**
```{r, echo = TRUE}
model2 <- lm(species_per_belt_loop ~ abs(belts))
summary(model2)

model3 <- lm(species_per_belt_loop ~ abs(belts) + I(abs(belts)^2))
summary(model3)

par(mfrow = c(1, 2))
plot(belts, species_per_belt_loop,
     xlab = "Latitude", ylab = "Species number", pch = 16)
plot(abs(belts), species_per_belt_loop,
     xlab = "Absolute latitude", ylab = "Species number",pch = 16)
abline(model2, col = "red", lwd = 2)
points(abs(belts), predict(model3), type = "l", col = "blue", lwd = 2)
AIC(model2, model3)
```

&nbsp;

Let's also assess model performance using the performance package to calculate commonly used indices, such as R^2^, AIC, RMSE, etc.  

```{r, echo = TRUE}
compare_performance(model3, model2)
```
