---
title: "Modern concepts and methods in Macroecology and Biogeography"
subtitle: "Day 5: Environmental drivers of palm diversity"
author: "Dylan Craven & Patrick Weigelt"
date: "May 16th, 2018"
output: 
  html_document: 
    theme: flatly
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Preparation:** Please open your RStudio project and download the data for today from [Stud-IP](https://www.studip.uni-goettingen.de/dispatch.php/file/details/b64b68db7a05f02fa854a63a01064f67?cid=95a0e5be352a34697791cddd2f4ddbf4) (log in first) and unzip them into your project folder '/data/. You can use `getwd()` to get your current working directory which is your project folder as long as you do not change it. Please install the following R-packages using `install.packages()` if not done yet:  

* rgdal  
* rgeos
* plyr  
* classInt
* colorRamps
* raster
* car


&nbsp;

### 1. Loading raster layers

**Clean up workspace**  
```{r}
rm(list=ls()) # You might want to remove all loaded objects first
```
&nbsp;

**Load R packages**  


```{r, include=FALSE}
library(plyr)
library(raster)
library(rgdal)
library(rgeos)
library(classInt)
library(colorRamps)
library(car)
```

```{r, eval=FALSE}
library(plyr)
library(raster)
library(rgdal)
library(rgeos)
library(classInt)
library(colorRamps)
library(car)
```
```{r, echo=FALSE}
if (file.exists("D:/rastertemp")){rasterOptions(tmpdir="D:/rastertemp/")} # raster saves large temporary files. You might want to change the location if you're running out of disk space
```

&nbsp;

**Load and extract environmental data from spatial layers**
For today's exercise, we again use the elevation, temperature ('bio1'), and precipitation ('bio12') raster layers we downloaded the first day.  

```{R, echo=TRUE}
# Import global digital elevation model at 30 arc seconds resolution
elev <- raster("Data/mn30_grd")

par(mfrow=c(1,1))
plot(elev) 
```
&nbsp;

```{R, echo=TRUE}
# Import climate data from Chelsa at 30 arc seconds resolution -->
bio1 <- raster("Data/CHELSA_bio10_1.tif") # temperature*10
bio12 <- raster("Data/CHELSA_bio10_12.tif") # precipitation
plot(bio1)
```
&nbsp;

<!-- rm(list=ls()) # remove everything  -->

<!-- library(plyr) -->
<!-- library(raster) -->
<!-- library(rgdal) -->
<!-- library(rgeos) -->
<!-- library(classInt) -->
<!-- library(colorRamps) -->
<!-- library(car) -->

<!-- if (file.exists("D:/rastertemp")){rasterOptions(tmpdir="D:/rastertemp/")} # raster saves large temporary files; You can define where  -->


<!-- grid <- readOGR("data/30min_grid_select50%.shp", integer64="allow.loss") -->

<!-- grid@proj4string <- CRS("+proj=longlat +ellps=WGS84 +no_defs") # remember to include coordinate system -->

<!-- # load raster library -->

<!-- # Import global digital elevation model at 30 arc seconds resolution -->
<!-- elev <- raster("data/mn30_grd") -->
<!-- par(mfrow=c(1,1)) -->
<!-- plot(elev) -->

<!-- # The raster layer is larger than it needs to be for Neotropical palm data. We can crop it based on the palm gridcell shapefile  -->
<!-- extent(grid) -->
<!-- elev <- crop(elev, extent(grid)) # crop raster layers to smaller extent -->
<!-- plot(elev) -->



<!-- # Import climate data from Chelsa at 30 arc seconds resolution -->
<!-- bio1 <- raster("data/CHELSA_bio10_1.tif") -->
<!-- bio1 <- crop(bio1, extent(grid)) # crop raster layers to smaller extent -->
<!-- bio12 <- raster("data/CHELSA_bio10_12.tif") -->
<!-- bio12 <- crop(bio12, extent(grid)) # crop raster layers to smaller extent -->

<!-- # To avoid broken numbers temperature has been multiplied by 10 in the 'bio1' raster layer.  -->
<!-- bio1 <- bio1/10 # Careful! this is computer intense and takes a long time -->

<!-- par(mfrow=c(1,2)) -->
<!-- plot(bio1) -->
<!-- plot(bio12) -->


<!-- ##################################################### -->
<!-- ### extract environmental information for the palm -->
<!-- ### grid polygons -->
<!-- par(mfrow=c(1,1)) -->
<!-- plot(bio1, ylim=c(-6,-2), xlim=c(-82,-76)) -->
<!-- plot(grid, add=TRUE) -->

<!-- ### Temperature -->
<!-- bio1_values <- extract(bio1,grid[c(1:10),]) -->
<!-- bio1_values -->
<!-- bio1_mean <- extract(bio1,grid[c(1:10),], fun=mean) -->
<!-- bio1_mean -->

<!-- head(grid@data) -->


<!-- # How long would it take for all gridcells? -->
<!-- nrow(grid) -->
<!-- nrow(grid)/10*system.time(bio1_mean <- extract(bio1,grid[c(1:10),], fun=mean))/60 -->
<!-- # 20 min? -->

<!-- ### faster work-around -->

<!-- # aggregate raster layer to the resolution of the palm grid -->
<!-- bio1_30min <- aggregate(bio1, fact=60, fun=mean, na.rm=TRUE) -->
<!-- dim(bio1) -->
<!-- extent(bio1) -->
<!-- dim(bio1_30min) -->
<!-- extent(bio1_30min) -->


<!-- # look at it -->
<!-- par(mfrow=c(1,2)) -->

<!-- plot(bio1, ylim=c(-6,-2), xlim=c(-82,-76)) -->
<!-- plot(grid, add=TRUE) -->

<!-- plot(bio1_30min, ylim=c(-6,-2), xlim=c(-82,-76)) -->
<!-- plot(grid, add=TRUE) -->

<!-- grid_centroids <- gCentroid(grid, byid=TRUE) -->
<!-- plot(grid_centroids, add=TRUE) -->

<!-- # extract values based on spatial points shapefile -->
<!-- temp_mean <- extract(bio1_30min,grid_centroids) -->
<!-- #temp_mean <- temp_mean/10 -->
<!-- temp_mean[1:10] -->

<!-- ### Precipitation -->
<!-- bio12_30min <- aggregate(bio12, fact=60, fun=mean, na.rm=TRUE) -->
<!-- prec_mean <- extract(bio12_30min,grid_centroids) -->
<!-- prec_mean[1:10] -->

<!-- ### Elevation range -->
<!-- elev_30min_min <- aggregate(elev, fact=60, fun=min, na.rm=TRUE) -->
<!-- elev_30min_max <- aggregate(elev, fact=60, fun=max, na.rm=TRUE) -->
<!-- elev_min <- extract(elev_30min_min,grid_centroids) -->
<!-- elev_max <- extract(elev_30min_max,grid_centroids) -->
<!-- elev_range <- elev_max - elev_min -->
<!-- elev_range[1:10] -->
<!-- hist(elev_range) -->





<!-- # Write extracted values into attribute table of gridcell shapefile -->
<!-- grid@data$temp <- temp_mean -->
<!-- grid@data$prec <- prec_mean -->
<!-- grid@data$elev <- elev_range -->

<!-- head(grid@data) -->

<!-- hist(grid@data$temp) -->
<!-- hist(grid@data$prec) -->
<!-- hist(grid@data$elev) -->

<!-- cor(grid@data[,c("temp","prec","elev")]) -->


<!-- # join environmental data and species numbers -->
<!-- names(grid@data)[1] <- "grid_id" -->


<!-- # read in RData files from hard drive -->
<!-- load("data/species_num.RData") -->

<!-- hist(species_num$spec_num) -->

<!-- #Exercise -->

<!-- #1) Combine the species numbers ('species_num') and the newly extracted environmental data in one table -->
<!-- species_num <- join(species_num, grid@data) -->

<!-- #2) Calculate 3 linear models of species richness with each environmental variable as predictor and compare their output and AIC values -->

<!-- #model -->
<!-- model_temp <- lm(species_num$spec_num ~ species_num$temp) -->
<!-- summary(model_temp) -->

<!-- model_prec <- lm(species_num$spec_num ~ species_num$prec) -->
<!-- summary(model_prec) -->

<!-- model_elev <- lm(species_num$spec_num ~ species_num$elev) -->
<!-- summary(model_elev) -->

<!-- AIC(model_temp,model_prec,model_elev) -->

<!-- #2) Make scatterplots for all three models and plot their regressions lines  -->
<!-- par(mfrow=c(1,3)) -->
<!-- plot(species_num$spec_num ~ species_num$temp) -->
<!-- abline(model_temp, col="red") -->
<!-- plot(species_num$spec_num ~ species_num$prec) -->
<!-- abline(model_prec, col="red") -->
<!-- plot(species_num$spec_num ~ species_num$elev) -->
<!-- abline(model_elev, col="red") -->



<!-- # multipredictor model -->
<!-- model_full <- lm(species_num$spec_num ~ species_num$temp + species_num$prec + species_num$elev) -->
<!-- summary(model_full) -->

<!-- model_temp_prec <- lm(species_num$spec_num ~ species_num$temp + species_num$prec) -->
<!-- model_temp_elev <- lm(species_num$spec_num ~ species_num$temp + species_num$elev) -->
<!-- model_elev_prec <- lm(species_num$spec_num ~ species_num$prec + species_num$elev) -->

<!-- AIC(model_full,model_temp_prec,model_temp_elev,model_elev_prec) -->

<!-- library(car) -->
<!-- crPlots(model_full) -->




<!-- #plot richness, richness predictions and residuals -->
<!-- species_num$pred <- predict(model_full) -->
<!-- species_num$resi <- residuals(model_full) -->


<!-- head(grid@data) -->
<!-- head(species_num) -->

<!-- grid@data <- join(grid@data, species_num[,c("grid_id","spec_num","pred","resi")], type="left", by="grid_id") -->

<!-- par(mfrow=c(1,3), mar=c(1,1,1.5,1)) -->
<!-- #Create color scheme -->
<!-- americas <- readOGR("data/americas.shp") -->

<!-- my.class.fr<-classIntervals(grid@data$spec_num, n=10, style="equal", dataPrecision=0) # bin data into n quantiles -->
<!-- my.pal<-matlab.like(10) -->
<!-- my.col.fr<-findColours(my.class.fr,my.pal) # ramp colors based on classInts -->

<!-- plot(grid, col=my.col.fr, border=NA, main="Palm species richness") -->
<!-- plot(americas, add=TRUE) -->

<!-- legend("bottomleft", # position -->
<!--        legend = names(attr(my.col.fr, "table")),  -->
<!--        title = "Species number", -->
<!--        fill = attr(my.col.fr, "palette"), -->
<!--        cex = 0.6, -->
<!--        bty = "n") # no box around it -->


<!-- my.class.fr<-classIntervals(grid@data$pred, n=10, style="equal", dataPrecision=0) # bin data into n quantiles -->
<!-- my.pal<-matlab.like(10) -->
<!-- my.col.fr<-findColours(my.class.fr,my.pal) # ramp colors based on classInts -->

<!-- plot(grid, col=my.col.fr, border=NA, main="Model predictions") -->
<!-- plot(americas, add=TRUE) -->

<!-- legend("bottomleft", # position -->
<!--        legend = names(attr(my.col.fr, "table")),  -->
<!--        title = "Species number", -->
<!--        fill = attr(my.col.fr, "palette"), -->
<!--        cex = 0.6, -->
<!--        bty = "n") # no box around it -->



<!-- my.class.fr<-classIntervals(grid@data$pred, n=10, style="equal", dataPrecision=0) # bin data into n quantiles -->
<!-- my.pal<-green2red(10) -->
<!-- my.col.fr<-findColours(my.class.fr,my.pal) # ramp colors based on classInts -->

<!-- plot(grid, col=my.col.fr, border=NA, main="Model residuals") -->
<!-- plot(americas, add=TRUE) -->

<!-- legend("bottomleft", # position -->
<!--        legend = names(attr(my.col.fr, "table")),  -->
<!--        title = "Species number", -->
<!--        fill = attr(my.col.fr, "palette"), -->
<!--        cex = 0.6, -->
<!--        bty = "n") # no box around it -->



<!-- #Excercise -->

<!-- #run a model of species richness in dependence on latititude + latitude² and compare the AIC to our best model with all environmental precitors -->

<!-- model_lat <- lm(species_num$spec_num ~ abs(species_num$Lat) + I(abs(species_num$spec_num)^2)) -->
<!-- summary(model_lat) -->
<!-- AIC(model_lat, model_full) -->


