---
title: "Modern concepts and methods in Macroecology and Biogeography"
subtitle: "Day 5: Environmental drivers of palm diversity"
author: "Dylan Craven & Patrick Weigelt"
date: "May 16th, 2018"
output: 
  html_document: 
    theme: flatly
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Preparation:** Today we will only use data already distributed during the last couple of weeks. If you don't have them all, please dowload from [Stud-IP](https://www.studip.uni-goettingen.de/) and unzip them into your project folder '/data/. You can use `getwd()` to get your current working directory which is your project folder as long as you do not change it. Please install the following R-packages using `install.packages()` if not done yet:  

* rgdal  
* rgeos
* plyr  
* classInt
* colorRamps
* raster
* car


&nbsp;

### 1. Loading and preparing environmental data from spatial raster layers

**Clean up workspace**  
```{r}
rm(list=ls()) # You might want to remove all loaded objects first
```
&nbsp;

**Load R packages**  


```{r, include=FALSE}
library(plyr)
library(raster)
library(rgdal)
library(rgeos)
library(classInt)
library(colorRamps)
library(car)
```

```{r, eval=FALSE}
library(plyr)
library(raster)
library(rgdal)
library(rgeos)
library(classInt)
library(colorRamps)
library(car)
```
```{r, echo=FALSE}
if (file.exists("D:/rastertemp")){rasterOptions(tmpdir="D:/rastertemp/")} # raster saves large temporary files. You might want to change the location if you're running out of disk space
```

&nbsp;

**Load environmental raster layers**  
For today's exercise, we again use the elevation, temperature ('bio1'), and precipitation ('bio12') raster layers we downloaded the first day.  

```{R, echo=TRUE}
# Import global digital elevation model at 30 arc seconds resolution
elev <- raster("data/mn30_grd")

par(mfrow=c(1,1))
plot(elev, main="Elevation a.s.l. (m)") 
```
&nbsp;

```{R, echo=TRUE}
# Import climate data from Chelsa at 30 arc seconds resolution -->
bio1 <- raster("data/CHELSA_bio10_1.tif") # temperature*10
bio12 <- raster("data/CHELSA_bio10_12.tif") # precipitation
plot(bio1, main="Annual mean temperature (10 x C°)")
```
&nbsp;


**Load gridcell shapefile** 
```{r, echo=TRUE}
grid <- readOGR("Data/30min_grid_select50%.shp", integer64="allow.loss") # read in polygon shapefile
grid@proj4string <- CRS("+proj=longlat +ellps=WGS84 +no_defs") # remember to include coordinate system
americas <- readOGR("data/americas.shp")
americas@proj4string <- CRS("+proj=longlat +ellps=WGS84 +no_defs")
plot(grid, border="darkgrey")
plot(americas, add=TRUE)
```

The raster layers have larger extents than  they need to have for our Neotropical palm data. We can crop them based on the palm gridcell shapefile
```{r, echo=TRUE, fig.height=3, fig.width=12}
extent(grid)
elev <- crop(elev, extent(grid)) # crop raster layers to smaller extent
bio1 <- crop(bio1, extent(grid))
bio12 <- crop(bio12, extent(grid))
par(mfrow=c(1,3))
plot(elev, main="Elevation a.s.l. (m)")
plot(bio1, main="Annual mean temperature (10 x C°)")
plot(bio12, main="Mean annual precipitation (mm)")
```
&nbsp;

To avoid broken numbers temperature has been multiplied by 10 in the 'bio1' raster layer. We can undo this by dividing by 10
```{r, echo=TRUE}
bio1 <- bio1/10 # Careful, This can take a while!
```
&nbsp;


### 2. Extracting environmental data for palm gridcells from raster layers
```{r, echo=TRUE, fig.height=3, fig.width=4.3}
par(mfrow=c(1,1))
plot(bio1, ylim=c(-6,-2), xlim=c(-82,-76), main="Elevation a.s.l. (m)")
plot(grid, add=TRUE)
```

**Extract values** 
```{r, echo=TRUE}
bio1_values <- extract(bio1,grid[c(1:10),])
str(bio1_values) #bio1_values is a list
bio1_values[[1]][1:10] #Use [[]] to index a list
mean(bio1_values[[1]])
bio1_mean <- sapply(bio1_values, mean) # apply a function to all elements of a list
bio1_mean
bio1_mean <- extract(bio1,grid[c(1:10),], fun=mean) # define the function for aggregation already in extract()
bio1_mean
```


How long would it take to extract mean values for all gridcells?
```{r, echo=TRUE}
nrow(grid) # number of gridcells
nrow(grid)/10*system.time(bio1_mean <- extract(bio1,grid[c(1:10),], fun=mean))/60
# 20 min?
```
&nbsp;


**Faster work-around: aggregating the raster layers**  
Aggregate raster layer to the resolution of the palm grid and convert grid to spatial points  
```{r, echo=TRUE}
bio1_30min <- aggregate(bio1, fact=60, fun=mean, na.rm=TRUE)
dim(bio1)
extent(bio1)
dim(bio1_30min)
extent(bio1_30min)

grid_centroids <- gCentroid(grid, byid=TRUE)
```
```{r, echo=TRUE, fig.height=3, fig.width=7.5}
par(mfrow=c(1,2))

plot(bio1, ylim=c(-6,-2), xlim=c(-82,-76), main="30 seconds resolution")
plot(grid, add=TRUE)

plot(bio1_30min, ylim=c(-6,-2), xlim=c(-82,-76), main="30 minutes resolution")
plot(grid, add=TRUE)
plot(grid_centroids, add=TRUE)
```


**Extract values based on spatial points and aggregated raster** 
```{r, echo=TRUE}
# Temperature
temp_mean <- extract(bio1_30min,grid_centroids) # extract values
temp_mean[1:10]

# Precipitation
bio12_30min <- aggregate(bio12, fact=60, fun=mean, na.rm=TRUE)
prec_mean <- extract(bio12_30min,grid_centroids)
prec_mean[1:10]

# Elevation range as a measure of topographic heterogeneity
elev_30min_min <- aggregate(elev, fact=60, fun=min, na.rm=TRUE) # minimum elevation
elev_30min_max <- aggregate(elev, fact=60, fun=max, na.rm=TRUE) # maximum elevation
elev_min <- extract(elev_30min_min,grid_centroids)
elev_max <- extract(elev_30min_max,grid_centroids)
elev_range <- elev_max - elev_min
elev_range[1:10]
```
&nbsp;

**Write extracted values into attribute table of gridcell shapefile**  
```{r, echo=TRUE, fig.height=4, fig.width=12}
grid@data$temp <- temp_mean
grid@data$prec <- prec_mean
grid@data$elev <- elev_range

head(grid@data)

cor(grid@data[,c("temp","prec","elev")])
par(mfrow=c(1,3))
hist(grid@data$temp, main="", xlab="Elevation a.s.l. (m)")
hist(grid@data$prec, main="", xlab="Annual mean temperature (C°)")
hist(grid@data$elev, main="", xlab="Mean annual precipitation (mm)")

save(grid, file="data/grid_env.RData")
```
&nbsp;

### 3. Modelling palm species richness  

**Join environmental data and species numbers**  
```{r, echo=TRUE, fig.height=4, fig.width=4}
load("data/grid_env.RData")
load("data/species_num.RData") # read in RData files of palm species richness from last week
hist(species_num$spec_num, main="", xlab="Palm species number")
```


#### Exercise  

1) Combine the species numbers ('species_num') and the newly extracted environmental data in one table.   
```{r, echo=FALSE}
names(grid@data)[1] <- "grid_id" # rename ID column in gridcell shapefile
species_num <- join(species_num, grid@data, by="grid_id")
```
&nbsp;

2) Calculate 3 linear models of species richness with each environmental variable as predictor and compare their output and AIC values.  
```{r, echo=FALSE}
model_elev <- lm(species_num$spec_num ~ species_num$elev)
summary(model_elev)

model_temp <- lm(species_num$spec_num ~ species_num$temp)
summary(model_temp)

model_prec <- lm(species_num$spec_num ~ species_num$prec)
summary(model_prec)


AIC(model_temp,model_prec,model_elev)
```
&nbsp;

3) Make scatterplots for all three models and plot their regression lines
```{r, echo=FALSE, fig.height=4, fig.width=12}
par(mfrow=c(1,3))
plot(species_num$spec_num ~ species_num$elev, xlab="Elevation a.s.l. (m)",  ylab="Species number")
abline(model_elev, col="red")
plot(species_num$spec_num ~ species_num$temp, xlab="Annual mean temperature (C°)",  ylab="Species number")
abline(model_temp, col="red")
plot(species_num$spec_num ~ species_num$prec, xlab="Mean annual precipitation (mm)",  ylab="Species number")
abline(model_prec, col="red")
```


**Multipredictor model**
```{r, echo=TRUE, fig.height=4, fig.width=12}
model_full <- lm(species_num$spec_num ~ species_num$temp + species_num$prec + species_num$elev)
summary(model_full)

model_temp_prec <- lm(species_num$spec_num ~ species_num$temp + species_num$prec)
model_temp_elev <- lm(species_num$spec_num ~ species_num$temp + species_num$elev)
model_elev_prec <- lm(species_num$spec_num ~ species_num$prec + species_num$elev)

AIC(model_full,model_temp_prec,model_temp_elev,model_elev_prec)

crPlots(model_full, layout=c(1, 3), ask=FALSE, main="Partial residual plots", smoother=FALSE, grid=FALSE, ylab="Partial residuals (species richness)")
```



**Plot species richness, model predictions and residuals**

Get predictions and get residuals and join them to gridcell shapefile
```{r, echo=TRUE, fig.height=4, fig.width=12}
species_num$pred <- predict(model_full)
species_num$resi <- residuals(model_full)
head(species_num)
head(grid@data)
grid@data <- join(grid@data, species_num[,c("grid_id","spec_num","pred","resi")], type="left", by="grid_id")

par(mfrow=c(1,3), mar=c(1,1,2,1))
#Create color scheme

my.class.fr<-classIntervals(grid@data$spec_num, n=10, style="equal", dataPrecision=0) # bin data into n quantiles
my.pal<-matlab.like(10)
my.col.fr<-findColours(my.class.fr,my.pal) # ramp colors based on classInts

plot(grid, col=my.col.fr, border=NA, main="Palm species richness")
plot(americas, add=TRUE)

legend("bottomleft", # position
       legend = names(attr(my.col.fr, "table")),
       title = "Species number",
       fill = attr(my.col.fr, "palette"),
       cex = 1,
       bty = "n") # no box around it


my.class.fr<-classIntervals(grid@data$pred, n=10, style="equal", dataPrecision=0) # bin data into n quantiles
my.pal<-matlab.like(10)
my.col.fr<-findColours(my.class.fr,my.pal) # ramp colors based on classInts

plot(grid, col=my.col.fr, border=NA, main="Model predictions")
plot(americas, add=TRUE)

legend("bottomleft", # position
       legend = names(attr(my.col.fr, "table")),
       title = "Species number",
       fill = attr(my.col.fr, "palette"),
       cex = 1,
       bty = "n") # no box around it



my.class.fr<-classIntervals(grid@data$pred, n=10, style="equal", dataPrecision=0) # bin data into n quantiles
my.pal<-green2red(10)
my.col.fr<-findColours(my.class.fr,my.pal) # ramp colors based on classInts

plot(grid, col=my.col.fr, border=NA, main="Model residuals")
plot(americas, add=TRUE)

legend("bottomleft", # position
       legend = names(attr(my.col.fr, "table")),
       title = "Species number",
       fill = attr(my.col.fr, "palette"),
       cex = 1,
       bty = "n") # no box around it
```

#### Exercise  

1) Run a model of species richness in dependence on latititude + latitude² and compare the AIC to our best model with all environmental precitors  
```{r, echo=FALSE}
model_lat <- lm(species_num$spec_num ~ abs(species_num$Lat) + I(abs(species_num$Lat)^2))
summary(model_lat)
AIC(model_lat, model_full)
```




