---
title: "Modern concepts and methods in Macroecology and Biogeography"
subtitle: "Day 5: Environmental drivers of neotropical palm diversity"
author: "Patrick Weigelt, Pierre Denelle & Holger Kreft"
date: "May 18th, 2022"
output: 
  html_document: 
    theme: flatly
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

&nbsp;
_[back to overview](index.html)_
&nbsp;
&nbsp;


**Preparation:** Open your RStudio project and download new data for today from [Stud-IP](https://studip.uni-goettingen.de/dispatch.php/course/files/index/74501b33f332a6d238443bfbd502cc6c?cid=1746ae3fe971afbf90b74e3d64389586) (log in first) and unzip them into your project folder '/data/. You can use `getwd()` to get your current working directory which is your project folder as long as you do not change it. Please install the following R-packages using `install.packages()` if not done yet:  

* sf
* raster
* dplyr
* RColorBrewer
* car
* ggplot2
* viridis
* cowplot

&nbsp;

If you want to visualize this tutorial in the viewer inside RStudio (to save space on your screen) run the following chunk of code:
```{r viewer, eval=FALSE}
install.packages("rstudioapi") # install an R-package required for this step
dir <- tempfile()
dir.create(dir)
download.file("https://gift.uni-goettingen.de/mcmmb/index.html", destfile = file.path(dir, "index.html"))
download.file("https://gift.uni-goettingen.de/mcmmb/Day5.html", destfile = file.path(dir, "Day5.html"))
htmlFile <- file.path(dir, "Day5.html")
rstudioapi::viewer(htmlFile)
```
Now you can conveniently copy code from the viewer into your script.  
&nbsp;


### 1. Loading and preparing environmental data from spatial raster layers

**Clean up workspace**  
```{r}
rm(list=ls()) # You might want to remove all loaded objects first
```
&nbsp;

**Load R packages**  


```{r, include=FALSE}
library(dplyr)
library(raster)
library(sf)
library(RColorBrewer)
library(car)
library(ggplot2)
library(viridis)
library(cowplot)
```

```{r, eval=FALSE}
library(dplyr)
library(raster)
library(sf)
library(RColorBrewer)
library(car)
library(ggplot2)
library(viridis)
library(cowplot)
```
```{r, echo=FALSE}
if (file.exists("D:/rastertemp")){rasterOptions(tmpdir="D:/rastertemp/")} # raster saves large temporary files. You might want to change the location if you're running out of disk space
```

&nbsp;

**Load environmental raster layers**  
For today's exercise, we again use the elevation, temperature ('bio1'), and precipitation ('bio12') raster layers we downloaded the first day.  

```{R, echo=TRUE}
# Import global digital elevation model at 30 arc seconds resolution
elev <- raster("data/mn30_grd")
plot(elev, main="Elevation a.s.l. (m)", col = terrain.colors(255)) 
```
&nbsp;

```{R, echo=TRUE}
# Import climate data from Chelsa at 30 arc seconds resolution -->
bio1 <- raster("data/CHELSA_bio10_01.tif") # temperature*10
bio12 <- raster("data/CHELSA_bio10_12.tif") # precipitation
plot(bio1, main="Annual mean temperature (10 x C°)", col = inferno(255))
```
&nbsp;


**Load gridcell shapefile** 
```{r, echo=TRUE}
# gridcells corresponding to palm distribution data and americas coastline
grid <- st_read("data/30min_grid_select50/30min_grid_select50%.shp")
americas <- st_read("data/americas/americas.shp")

# remember to define the Coordinate Reference System (CRS)
st_crs(grid) <- "+proj=longlat +ellps=WGS84 +no_defs"
st_crs(americas) <- "+proj=longlat +ellps=WGS84 +no_defs"

ggplot(grid) +
  geom_sf() +
  geom_sf(data = americas, fill = NA, color = "black") +
  theme_void() + ylim(-40, 40)
```

The raster layers have larger extents than they need to have for our Neotropical palm data. We can crop them based on the palm gridcell shapefile:
```{r, echo=TRUE, fig.height=3, fig.width=12}
extent(grid)
elev <- crop(elev, extent(grid)) # crop raster layers to smaller extent
bio1 <- crop(bio1, extent(grid))
bio12 <- crop(bio12, extent(grid))
elev <- elev * bio1 / bio1

par(mfrow=c(1,3))
plot(elev, main="Elevation a.s.l. (m)", col = terrain.colors(255))
plot(bio1, main="Annual mean temperature (10 x C°)", col = inferno(255))
plot(bio12, main="Mean annual precipitation (mm)", col = colorRampPalette(brewer.pal(9,"Blues"))(255))
```
&nbsp;

To avoid broken numbers temperature has been multiplied by 10 in the 'bio1' raster layer. We can undo this by dividing by 10
```{r, echo=TRUE}
bio1 <- bio1/10 # Careful, This can take a while!
```
&nbsp;


### 2. Extracting environmental data for palm gridcells from raster layers
```{r, echo=TRUE, fig.height=3, fig.width=4.3}
par(mfrow=c(1,1))
plot(elev, ylim=c(-6,-2), xlim=c(-82,-76), main="Elevation a.s.l. (m)", col = terrain.colors(255))
plot(st_geometry(grid), add=TRUE)
```

**Extract values** 
```{r, echo=TRUE}
bio1_values <- extract(bio1,grid[c(1:10),])
str(bio1_values) #bio1_values is a list
bio1_values[[1]][1:10] #Use [[]] to index a list
mean(bio1_values[[1]])
bio1_mean <- sapply(bio1_values, mean) # apply a function to all elements of a list
bio1_mean
bio1_mean <- extract(bio1,grid[c(1:10),], fun=mean) # define the function for aggregation already in extract()
bio1_mean
```


How long would it take to extract mean values for all gridcells?
```{r, echo=TRUE}
nrow(grid) # number of gridcells
nrow(grid)/10*system.time(bio1_mean <- extract(bio1,grid[c(1:10),], fun=mean))/60
# 20 min?
```
Note: There are faster packages for dealing with raster layers available now (_velox_, _fasterRaster_). Here, we want to make use of a workaround that also helps us top understand the data better.   
&nbsp;


**Faster work-around: aggregating the raster layers**  
Aggregate raster layer to the resolution of the palm grid and convert grid to spatial points  
```{r, echo=TRUE}
bio1_30min <- aggregate(bio1, fact=60, fun=mean, na.rm=TRUE)
dim(bio1)
extent(bio1)
dim(bio1_30min)
extent(bio1_30min)

grid_centroids <- st_centroid(grid)
```
```{r, echo=TRUE, fig.height=3, fig.width=7.5}
par(mfrow=c(1,2))

plot(bio1, ylim=c(-6,-2), xlim=c(-82,-76), col = inferno(255), main="30 seconds resolution")
plot(st_geometry(grid), add=TRUE)

plot(bio1_30min, ylim=c(-6,-2), xlim=c(-82,-76), col = inferno(255), main="30 minutes resolution")
plot(st_geometry(grid), add=TRUE)
plot(st_geometry(grid_centroids), add=TRUE, pch = 3)
```


**Extract values based on spatial points and aggregated raster** 
```{r, echo=TRUE}
# Temperature
temp_mean <- extract(bio1_30min,grid_centroids) # extract values
temp_mean[1:10]

# Precipitation
bio12_30min <- aggregate(bio12, fact=60, fun=mean, na.rm=TRUE)
prec_mean <- extract(bio12_30min,grid_centroids)
prec_mean[1:10]

# Elevation range as a measure of topographic heterogeneity
elev_30min_min <- aggregate(elev, fact=60, fun=min, na.rm=TRUE) # minimum elevation
elev_30min_max <- aggregate(elev, fact=60, fun=max, na.rm=TRUE) # maximum elevation
elev_min <- extract(elev_30min_min, grid_centroids)
elev_max <- extract(elev_30min_max, grid_centroids)
elev_range <- elev_max - elev_min
elev_range[1:10]
```
&nbsp;

**Write extracted values into gridcell shapefile**  
```{r, echo=TRUE, fig.height=4, fig.width=12}
grid$temp <- temp_mean
grid$prec <- prec_mean
grid$elev <- elev_range

head(grid)

cor(st_drop_geometry(grid[,c("temp","prec","elev")]))

plot_grid(nrow = 1, ncol = 3,
ggplot(grid, aes(temp)) +
  geom_histogram(fill = "grey", color = "black", bins = 30) + 
  labs(x = "Annual mean temperature (C°)", y = "Count") +
  theme_bw(),
ggplot(grid, aes(prec)) +
  geom_histogram(fill = "grey", color = "black", bins = 30) + 
  labs(x = "Mean annual precipitation (mm)", y = "Count") +
  theme_bw(),
ggplot(grid, aes(elev)) +
  geom_histogram(fill = "grey", color = "black", bins = 30) + 
  labs(x = "Elevation a.s.l. (m)", y = "Count") +
  theme_bw()
)
# save(grid, file="Data/grid_env.RData")
```
&nbsp;

### 3. Modelling palm species richness  

**Join environmental data and species numbers**  
```{r, echo=TRUE, fig.height=4, fig.width=4}
load("data/grid_env.RData")
load("data/species_num.RData") # read in RData files of palm species richness from last week

ggplot(species_num, aes(spec_num)) +
  geom_histogram(fill = "grey", color = "black", bins = 30) + 
  labs(x = "Species number", y = "Count") +
  theme_bw()
```


#### Exercise  

1) Join the species numbers ('species_num') to the grid shapefile including the newly extracted environmental data.   
```{r, echo=FALSE}
names(grid)[1] <- "grid_id" # rename ID column in gridcell shapefile
grid <- left_join(grid, species_num, by="grid_id")
head(grid)
```
&nbsp;

2) Calculate 3 linear models of species richness with each environmental variable as predictor and compare their output and AIC values.  
```{r, echo=FALSE}
model_elev <- lm(grid$spec_num ~ grid$elev)
summary(model_elev)

model_temp <- lm(grid$spec_num ~ grid$temp)
summary(model_temp)

model_prec <- lm(grid$spec_num ~ grid$prec)
summary(model_prec)


AIC(model_temp,model_prec,model_elev)
```
&nbsp;

3) Make scatterplots for all three models and plot their regression lines
```{r, echo=FALSE, fig.height=4, fig.width=12}
# par(mfrow=c(1,3))
# plot(grid$spec_num ~ grid$elev, xlab="Elevation a.s.l. (m)",  ylab="Species number")
# abline(model_elev, col="red")
# plot(grid$spec_num ~ grid$temp, xlab="Annual mean temperature (C°)",  ylab="Species number")
# abline(model_temp, col="red")
# plot(grid$spec_num ~ grid$prec, xlab="Mean annual precipitation (mm)",  ylab="Species number")
# abline(model_prec, col="red")

plot_grid(nrow = 1, ncol = 3,
ggplot(grid, aes(elev, spec_num)) +
  geom_point(color = "black", alpha = 0.3) +
  stat_smooth(method = "lm", formula = y ~ x, color = "red") +
  labs(x = "Elevation a.s.l. (m)", y = "Species number") +
  theme_bw(),
ggplot(grid, aes(temp, spec_num)) +
  geom_point(color = "black", alpha = 0.3) +
  stat_smooth(method = "lm", formula = y ~ x, color = "red") +
  labs(x = "Annual mean temperature (C°)", y = "Species number") +
  theme_bw(),
ggplot(grid, aes(prec, spec_num)) +
  geom_point(color = "black", alpha = 0.3) +
  stat_smooth(method = "lm", formula = y ~ x, color = "red") +
  labs(x = "Mean annual precipitation (mm)", y = "Species number") +
  theme_bw())

```


**Multipredictor model**
```{r, echo=TRUE, fig.height=4, fig.width=12}
model_full <- lm(grid$spec_num ~ grid$temp + grid$prec + grid$elev)
summary(model_full)

model_temp_prec <- lm(grid$spec_num ~ grid$temp + grid$prec)
model_temp_elev <- lm(grid$spec_num ~ grid$temp + grid$elev)
model_elev_prec <- lm(grid$spec_num ~ grid$prec + grid$elev)

AIC(model_full,model_temp_prec,model_temp_elev,model_elev_prec)

crPlots(model_full, layout=c(1, 3), ask=FALSE, main="Partial residual plots", grid=FALSE, ylab="Partial residuals (species richness)")
```



**Plot species richness, model predictions and residuals**

Get predictions and get residuals and join them to gridcell shapefile
```{r, echo=TRUE, fig.height=4, fig.width=12}
grid$pred <- predict(model_full)
grid$resi <- residuals(model_full)
head(grid)


lmt <- ceiling(10*max(abs(range(grid$resi))))/10

plot_grid(nrow = 1, ncol = 3,
  ggplot(grid) +
    geom_sf(color = NA, aes(fill = spec_num)) +
    geom_sf(data = americas, fill = NA, color = "black") +
    labs(title = "Palm species richness") +
    scale_fill_viridis_c("Species number") +
    theme_void() + ylim(-40, 40),
  ggplot(grid) +
    geom_sf(color = NA, aes(fill = pred)) +
    geom_sf(data = americas, fill = NA, color = "black") +
    labs(title = "Model predictions") +
    scale_fill_viridis_c("Species number") +
    theme_void() + ylim(-40, 40),
  ggplot(grid) +
    geom_sf(color = NA, aes(fill = resi)) +
    geom_sf(data = americas, fill = NA, color = "black") +
    labs(title = "Model residuals") +
    # scale_fill_distiller("Species number", palette="PiYG") +
    scale_fill_stepsn(colors=brewer.pal(10, "PiYG"),
             n.breaks=10, limits=c(-lmt,lmt), show.limits=T) +
    theme_void() + ylim(-40, 40))
```

#### Exercise  

1) Run a model of species richness in dependence on latitude + latitude² and compare the AIC to our best model with all environmental predictors  
```{r, echo=FALSE}
model_lat <- lm(grid$spec_num ~ abs(grid$Lat) + I(abs(grid$Lat)^2))
summary(model_lat)
AIC(model_lat, model_full)

ggplot(grid, aes(abs(Lat), spec_num)) +
  geom_point(color = "black", alpha = 0.3) +
  stat_smooth(method = "lm", formula = y ~ x + I(x^2), color = "red") +
  labs(x = "Latitude", y = "Species number") +
  theme_bw()
```




