---
title: "Modern concepts and methods in Macroecology and Biogeography"
subtitle: "Day 6: Island Biogeography"
author: "Dylan Craven & Patrick Weigelt"
date: "May 27th, 2019"
output: 
  html_document: 
    fig_height: 7
    theme: flatly
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

&nbsp;
_[back to overview](index.html)_
&nbsp;
&nbsp;


**Preparation:** Please open your RStudio project and download all new data ('islands.csv') for today from [Stud-IP](https://studip.uni-goettingen.de/dispatch.php/course/files?cid=a16be33e4d2eb3c1c82a459c44495012) (log in first) and copy it into your project folder '/Data/'. You can use `getwd()` to locate your current working directory, which should be your project folder. Please install the following R-packages using `install.packages()`:  

* maptools  
* maps  
* maptools  
* classInt  
* psych
* car


&nbsp;

**Load R packages & island data set**  


```{r, include=FALSE}
library(maptools)
library(maps)
library(maptools)
library(classInt)
data(worldMapEnv)
library(psych)
library(car)
```

```{r, eval=FALSE}
library(maptools)
library(maps)
library(maptools)
library(classInt)
data(worldMapEnv)
library(psych)
library(car)
```
&nbsp;

```{r}
islands <- read.csv("Data/islands.csv")
```

### 1. Data exploration: spatial patterns of species diversity \& co-variation among biogeographic and environmental variables   


**Create a colour gradient for species richness**  
```{r, echo=TRUE}

islands <- islands[order(islands$spec_num),] # arrange islands by species richness (low to high)

my.class.fr.S<-classIntervals(islands$spec_num,n=9, style="fixed", fixedBreaks=c(0,25,50,100,200,400,800,1600,3200,6400)) # bin data into n quantiles
my.pal<-c("midnightblue","dodgerblue4","dodgerblue2","chartreuse4","gold","orangered1","darkred")
my.col.fr.S<-findColours(my.class.fr.S,my.pal) # ramp colors based on classInts
my.col.fr.S[islands$spec_num < 1] <- "#666666" # assign grey for zero species

# specify dimensions of plot for saving
if(.Platform$OS.type == "windows"){ 
  windows(width=20*0.3937008, height=12*0.3937008)
}else{
  X11(width=20*0.3937008, height=12*0.3937008,type="cairo")
  }
# quartz(width=20*0.3937008, height=12*0.3937008) # for Mac 

map('world', fill = FALSE)
title(main = "Angiosperm species richness")
points(islands$longitude, islands$latitude, pch=ifelse(islands$spec_num >= 1,1,4), cex = (log10(islands$spec_num+1)/2.6)^2+1, col = my.col.fr.S, lwd=2)

if(.Platform$OS.type == "windows"){ 
  savePlot("Data/islands.pdf", type = "pdf")
}else{
  dev.copy2pdf(file = "Data/islands.pdf")
}

#quartz.save("Data/islands.pdf", type = "pdf") # for Mac 
dev.off()
```
&nbsp;


**Correlations among biogeographic \& environmental variables**

_Meta-data_    
1. **latitude**: Latitude (in decimal degrees)  
2. **area**: Island area (in km^2)    
3. **dist**: Distance to the nearest continent (in km)   
4. **age_Ma**: Island age (in millions of years)    
5. **max_elev**: Maximum elevation (m)     
6. **mean_temp**: Mean annual temperature (degrees Celsius)    
7. **mean_prec**: Mean annual precipitation (mm yr^-1)  
8. **mean_homogeneity**: Habitat heterogeneity (similarity of vegetation index with surrounding pixels)   


```{r,echo=TRUE}
pairs.panels(islands[,c(4:7,11:14)], density=F,ellipses=F, hist.col = "white")
```
&nbsp;

### 2. The species area relationship (SAR)  

_Species Richness ~ Area_  


**Linear model**  
```{r, echo=TRUE}

SAR0 <- lm(spec_num ~ area, data=islands)
summary(SAR0)

plot(spec_num~area,data=islands)
regLine(SAR0, col="darkgreen")
```
&nbsp;

**Non-linear model**

_Species richness ~ c + Area^z^_  

```{r, echo=TRUE}
SAR1 <- nls(spec_num ~ c + area^z, data = islands, start = list(c = 0, z = 1))
summary(SAR1)

newdata <- data.frame(area=seq(min(islands$area),max(islands$area),length.out = 100))
predicted_richness_SAR1 <- predict(SAR1, newdata = newdata)

plot(islands$area, islands$spec_num, xlab = "Island area (km)", ylab="Species number", main="Species-area relationship")
points(newdata$area,predicted_richness_SAR1, type="l", col="darkred", lwd=2)

```
&nbsp;


### Exercise I  
1. Make histograms of species richness and area  
2. Apply a log10 transformation to species richness and area, make new columns of the transformed variables, and plot histograms of them  
3. Fit a linear model to predict species richness using area 
4. Make a plot of the results  

&nbsp;


**Exercise I solutions: **  

_Make histograms of species richness and area_    

```{r, echo=FALSE}
par(mfrow=c(1,2))
# 1. look at the histograms of species richness and area
hist(islands$area)
hist(islands$spec_num)
```
&nbsp;

_Apply a log10 transformation to species richness and area, make new columns of the transformed variables, and plot histograms of them_
_&_
_Fit a linear model to predict species richness using area_

```{r, echo=FALSE}
islands$log_area <- log10(islands$area + 1)
islands$log_spec_num <- log10(islands$spec_num + 1)

par(mfrow=c(1,2))
hist(islands$log_area)
hist(islands$log_spec_num)

SAR2 <- lm(log_spec_num ~ log_area, data = islands)
summary(SAR2)

par(mfrow=c(1,1))
plot(log_spec_num~log_area, data=islands)
regLine(SAR2,col="darkgreen")
```


**Check model assumptions**  

```{r, echo=FALSE}
par(mfrow=c(1,2))
hist(residuals(SAR0))
hist(residuals(SAR2))
```

&nbsp;

**Poisson generalised linear model (GLM)** 

```{r, echo=TRUE}
SAR3 <- glm(spec_num ~ log_area, data=islands, family="poisson")

# Plot Poisson GLM with log-log model
plot(islands$log_area, islands$log_spec_num, xlab = "log10 island area (log10 km)", ylab="log10 species number", main="log-log Species-area relationship")
regLine(SAR2)

newdata_SAR3 <- data.frame(log_area=seq(min(islands$log_area),max(islands$log_area),length.out = 100))
predicted_richness_SAR3 <- predict(SAR3, newdata = newdata_SAR3, type="response")
points(newdata_SAR3$log_area,log10(predicted_richness_SAR3+1), type="l", col="purple", lwd=2)

#Now plot all models together 
plot(islands$area, islands$spec_num, xlab = "Island area (km?)", ylab="Species number", main="Species-area relationship")

# plot the simple linear model
regLine(SAR0, col="darkgreen")

# plot the non-linear power law model
points(newdata$area,predicted_richness_SAR1, type="l", col="darkred", lwd=2)

# plot the log-log space model in untransformed space
newdata_SAR2 <- data.frame(log_area=log10(seq(min(islands$area),max(islands$area),length.out = 100)+1))
predicted_richness_SAR2 <- predict(SAR2, newdata = newdata_SAR2)
points((10^newdata_SAR2$log_area)-1,(10^predicted_richness_SAR2)-1, type="l", col="purple", lwd=2)
points(newdata$area,(10^predicted_richness_SAR2)-1, type="l", col="darkgreen", lwd=2, lty=2)

# plot the Poisson model in untransformed space
predicted_richness_SAR3 <- predict(SAR3, newdata = newdata_SAR3, type="response")
points((10^newdata_SAR3$log_area)-1,predicted_richness_SAR3, type="l", col="darkblue", lwd=2)
```
&nbsp;

**How do these models compare?**  
  
```{r}

AIC(SAR0, SAR1, SAR2, SAR3)
```
&nbsp;

_Note_: You cannot compare AIC values among models with different response variables. But if models have the same response variable and *different* distribution families,
they can be compared used AIC (or equivalents).  

&nbsp;


### 3. Multi-predictor model of island biodiversity


### Exercise II    

1.  Apply a log10 transformation to area, elevation, precipitation, and temperature following:  
_Kreft, H. et al. 2008. Global diversity of island floras from a macroecological perspective. Ecology Letters 11:116-127._  
 
2. Simplify the geology column; change all values that are **floor** or **volcanic** to **oceanic**  

3. Fit a multiple linear regression to predict biodiversity patterns on islands using the following biogeographic and environmental
variables: island area, distance, temperature, precipitation, and island geology.  

4. Use manual backward selection to specify the best model  

**Hints**    
1. Remove missing observations from your data  
2. Use 'drop1(mod, test="F")'    
3. Use 'drop1(update(Multi.mod,~ . -log_max_elev), test = "F")'    

&nbsp;

**Solutions to Exercise II**

_Transform your data_
```{r, echo=FALSE}
islands$log_area<-log10(islands$area)
islands$log_spec_num<-log10(islands$spec_num+1)

islands$log_max_elev <- log10(islands$max_elev + 1)
islands$log_mean_prec <- log10(islands$mean_prec)
islands$log_mean_temp <- log10(islands$mean_temp+5)

```

&nbsp;

_Simplify the geology column_
```{r,echo=FALSE}
islands$geology<-as.character(islands$geology)
islands$geology[which(islands$geology == "floor" | islands$geology == "volcanic")] <- "oceanic"
islands$geology[which(islands$geology %in% c("floor" , "volcanic"))] <- "oceanic"

islands$geology[which(islands$geology == "floor")] <- "oceanic"
islands$geology[which(islands$geology == "volcanic")] <- "oceanic"

summary(as.factor(islands$geology))
```
&nbsp;

_Fit multiple regression model_  

```{r,echo=FALSE}

new_islands<-islands[,c(6,8,15:19)]
new_islands<-na.omit(new_islands) # remove observations with NAs

Multi.mod <- lm(log_spec_num ~ log_area + dist + log_max_elev+ log_mean_temp + log_mean_prec + geology, data= new_islands)

summary(Multi.mod)

crPlots(Multi.mod)
```
&nbsp;

_Can we improve predictions of species richness by removing individual model terms?_  

```{r,echo=FALSE}

Multi.mod1<-update(Multi.mod,~ . -log_max_elev) 
AIC(Multi.mod,Multi.mod1)
```
&nbsp;


### 4. The General dynamic model of island biogeography (GDM)  

```{r, echo=TRUE}

# Assign archipelago names for 4 major archipelagos
islands$archipelago <- NA
islands$archipelago[which(islands$entity_ID %in% c(169:188))] <- "Hawaii"
islands$archipelago[which(islands$entity_ID %in% c(145:151))] <- "Canaries"
islands$archipelago[which(islands$entity_ID %in% c(154:168))] <- "Galapagos"
islands$archipelago[which(islands$entity_ID %in% c(131:139))] <- "Azores"
islands_gdm <- islands[!is.na(islands$archipelago),]
```
&nbsp;

**The General dynamic model of island biogeography (GDM) applied: just Hawaii (part I)**

_Species = Area x Time x Time^2^_    


```{r, echo=TRUE}
Hawaii <- islands_gdm[which(islands_gdm$archipelago == "Hawaii"),] # subset data

# The ATT^2 model
gdm1 <- lm(spec_num ~ log_area + age_Ma + I(age_Ma^2), data=Hawaii)
summary(gdm1)
crPlots(gdm1)
```
&nbsp;

**The General dynamic model of island biogeography (GDM) applied: just Hawaii (part II)**

_Species = log(Area) x log(Time) x log(Time)^2^_    

```{r, echo=TRUE}
Hawaii$log_age_Ma <- log10(Hawaii$age_Ma)
gdm2 <- lm(spec_num ~ log_area + log_age_Ma + I(log_age_Ma^2), data=Hawaii)
summary(gdm2)
AIC(gdm1,gdm2)

crPlots(gdm1)
```
&nbsp;
