---
title: "Modern concepts and methods in Macroecology and Biogeography"
subtitle: "Day 6: Island Biogeography"
author: "Patrick Weigelt, Pierre Denelle & Holger Kreft"
date: "May 22nd, 2024"
output: 
  html_document: 
    fig_height: 7
    theme: flatly
    toc: yes
    toc_depth: 3
editor_options: 
 chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

&nbsp;
_[back to overview](index.html)_
&nbsp;
&nbsp;

**Preparation:** Open your RStudio project and download new data for today from [Stud-IP](https://studip.uni-goettingen.de/dispatch.php/course/files/index/74501b33f332a6d238443bfbd502cc6c?cid=1746ae3fe971afbf90b74e3d64389586) (log in first) and unzip them into your project folder '/data/. You can use `getwd()` to locate your current working directory, which should be your project folder. Please install the following R-packages using `install.packages()`:  

* maps  
* classInt  
* psych
* car
* lme4
* jtools
* performance
* ggplot2
* sf
* GGally
* rnaturalearth
* rnaturalearthdata
* ggpmisc
* cowplot
* dplyr
* tidyr
* GIFT

&nbsp;

If you want to visualize this tutorial in the viewer inside RStudio (to save space on your screen) run the following chunk of code:
```{r viewer, eval=FALSE}
install.packages("rstudioapi") # install an R-package required for this step
dir <- tempfile()
dir.create(dir)
download.file("https://gift.uni-goettingen.de/mcmmb/index.html", destfile = file.path(dir, "index.html"))
download.file("https://gift.uni-goettingen.de/mcmmb/Day6.html", destfile = file.path(dir, "Day6.html"))
htmlFile <- file.path(dir, "Day6.html")
rstudioapi::viewer(htmlFile)
```
Now you can conveniently copy code from the viewer into your script.  
&nbsp;

**Load R packages & island data set**  

```{r, include=FALSE}
library(maps)
library(classInt)
data(worldMapEnv)
library(psych)
library(car)
library(lme4)
library(jtools)
library(performance)
library(ggplot2)
library(GGally)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggpmisc)
library(cowplot)
library(dplyr)
library(tidyr)
library(GIFT)
```

```{r, eval=FALSE}
library(maps)
library(classInt)
data(worldMapEnv)
library(psych)
library(car)
library(lme4)
library(jtools)
library(performance)
library(ggplot2)
library(GGally)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggpmisc)
library(cowplot)
library(dplyr)
library(tidyr)
library(GIFT)
```
&nbsp;

# 1. Data extraction & exploration  

## 1.1. Data extraction from GIFT database

We will download island-level angiosperm species richness and environmental data from the  [GIFT](https://gift.uni-goettingen.de/home) database ([Weigelt et al. 2020](https://onlinelibrary.wiley.com/doi/10.1111/jbi.13623)) using the [GIFT R-package](https://biogeomacro.github.io/GIFT/).


```{r}
# Islands in GIFT
islands <- GIFT_regions(GIFT_version = "2.1")
islands <- islands[which(islands$suit_geo == 1 & islands$entity_class == "Island"), 
                   c("entity_ID", "geo_entity")]

# Geographic characteristics
islands_geo <- GIFT_env(entity_ID = islands$entity_ID,
                        miscellaneous = c("longitude","latitude","area","dist"),
                        GIFT_version = "2.1")
islands_geo$geo_entity <- NULL

islands <- inner_join(islands, islands_geo, by="entity_ID")

# Island geology
islands_geology <- read.csv("data/islands_geology.csv")

islands <- inner_join(islands, islands_geology, by="entity_ID")

# Angiosperm species richness
islands_richness <- GIFT_richness(taxon_name = "Angiospermae", GIFT_version = "2.1")
islands_richness <- islands_richness[!is.na(islands_richness$native),
                                     c("entity_ID","native","endemic_min")]
names(islands_richness) <- c("entity_ID","spec_num","end_num")

islands <- inner_join(islands, islands_richness, by="entity_ID")

# Environmental data
islands_env <- GIFT_env(entity_ID = islands$entity_ID,
                        rasterlayer = c("mx30_grd", "CHELSA_bio10_1", 
                                       "CHELSA_bio10_12", "Homogeneity_01_05_1km_uint16"),
                        sumstat = list("max","mean","mean","mean"),
                        GIFT_version = "2.1")

names(islands_env) <- c("entity_ID", "geo_entity", "max_elev", "mean_temp", "mean_prec", "mean_homogeneity")
islands_env$geo_entity <- NULL

islands <- inner_join(islands, islands_env, by="entity_ID")
str(islands)

# write.csv(islands, "data/islands.csv", row.names = FALSE)
# islands <- read.csv("data/islands.csv")
```


## 1.2. Map species richness

With ggplot2

```{r}
# World coastlines
world <- ne_coastline(scale = "medium", returnclass = "sf")
class(world)

# Other projections
eckertIV <- "+proj=eck4 +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs" 
mollweide <- "+proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs "
# https://semba-blog.netlify.app/01/26/2020/world-map-and-map-projections/

# Links regarding the choice of projections
# http://www.geography.hunter.cuny.edu/~jochen/gtech361/lectures/lecture04/concepts/Map%20coordinate%20systems/Map%20projections%20and%20distortion.htm
# http://www.geo.hunter.cuny.edu/~jochen/gtech201/lectures/lec6concepts/map%20coordinate%20systems/how%20to%20choose%20a%20projection.htm

plot_grid(
  ggplot(world) +
    geom_sf() +
    labs(title = "",
         subtitle = "WGS84 projection") +
    theme_bw() +
    theme(panel.border = element_blank()),
  ggplot(world) +
    geom_sf() +
    coord_sf(crs = eckertIV) +
    labs(title = "World map",
         subtitle = "EckertIV projection") +
    theme_bw() +
    theme(panel.border = element_blank()),
  ggplot(world) +
    geom_sf() +
    coord_sf(crs = mollweide) +
    labs(title = "",
         subtitle = "Mollweide projection") +
    theme_bw() +
    theme(panel.border = element_blank()),
  nrow = 2)
```

To combine the islands data.frame with the world map, we first need to convert it to an *sf* object.

```{r}
# Convert islands to sf object
islands_sf <- st_as_sf(islands, coords = c("longitude", "latitude"),
                       crs = "+proj=longlat +ellps=WGS84 +no_defs")

# Map
ggplot(world) +
  geom_sf() +
  geom_sf(data = islands_sf, shape = 1, stroke = 1.5,
          aes(size = spec_num, color = spec_num)) +
  scale_color_viridis_c("Species richness", trans = "log", guide = "legend") +
  scale_size_continuous("Species richness", trans = "log") +
  coord_sf(crs = mollweide) + # coord_sf(crs = eckertIV) +
  labs(title = "Angiosperm species richness",
       subtitle = "Mollweide projection") +
  theme_bw()

# https://community.rstudio.com/t/ggplot2-is-it-possible-to-combine-color-fill-and-size-legends/17072
```

*Bonus: custom discrete color scale*
<details><summary>Bonus</summary>
```{r, echo=TRUE, eval = FALSE}
# Arrange islands by species richness (low to high)
islands <- islands[order(islands$spec_num), ]

# create colour gradient
my.class.fr.S <- classIntervals(
  islands$spec_num, n = 9, style = "fixed",
  fixedBreaks = c(0, 25, 50, 100, 200, 400, 800, 1600, 3200, 6400)) # bin data into n quantiles
my.pal <- c("midnightblue", "dodgerblue4", "dodgerblue2", "chartreuse4",
            "gold", "orangered1", "darkred")
my.col.fr.S <- findColours(my.class.fr.S, my.pal) # ramp colors based on classInts
my.col.fr.S[islands$spec_num < 1] <- "#666666" # assign grey for zero species

# specify dimensions of plot for saving
if(.Platform$OS.type == "windows"){ 
  windows(width=20*0.3937008, height=12*0.3937008)
}else{
  X11(width=20*0.3937008, height=12*0.3937008,type="cairo")
}
# quartz(width=20*0.3937008, height=12*0.3937008) # for Mac 

map("world", fill = FALSE)
title(main = "Angiosperm species richness")
points(islands$longitude, islands$latitude,
       pch=ifelse(islands$spec_num >= 1,1,4),
       cex = (log10(islands$spec_num+1)/2.6)^2+1, col = my.col.fr.S, lwd=2)

if(.Platform$OS.type == "windows"){ 
  savePlot("data/islands.pdf", type = "pdf")
}else{
  dev.copy2pdf(file = "data/islands.pdf")
}

# quartz.save("Data/islands.pdf", type = "pdf") # for Mac 
dev.off()
```
&nbsp;
</details>

## 1.3. Correlations among biogeographic \& environmental variables

_Meta-data_    
1. **latitude**: Latitude (in decimal degrees)  
2. **area**: Island area (in km^2^)    
3. **dist**: Distance to the nearest continent (in km)   
4. **age_MA**: Island geological age (million years)     
5. **max_elev**: Maximum elevation (m)     
6. **mean_temp**: Mean annual temperature (degrees Celsius)    
7. **mean_prec**: Mean annual precipitation (mm yr^-1^)  
8. **mean_homogeneity**: Habitat heterogeneity (similarity of vegetation index with surrounding pixels)   

```{r, echo = TRUE, eval = FALSE}
abiotic <- c("latitude", "area", "dist", "age_MA", "max_elev",
             "mean_temp", "mean_prec", "mean_homogeneity")

ggpairs(islands[, abiotic],
        upper = list(continuous = wrap(ggally_cor, digits = 1))) +
  theme_bw()

# Other way to have a pair plot:
# pairs.panels(islands[, c(4:7, 11:14)],
#              density = FALSE, ellipses = FALSE,
#              hist.col = "white")
```

```{r, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE}
abiotic <- c("latitude", "area", "dist", "age_MA", "max_elev",
             "mean_temp", "mean_prec", "mean_homogeneity")

suppressWarnings(
  ggpairs(islands[, abiotic],
          upper = list(continuous = wrap(ggally_cor, digits = 1))) +
    theme_bw()
)
```

&nbsp;

# 2. The species area relationship (SAR)  
We here aim at linking, using the best model, island area and plant species richness.  

## a) Linear model  
_Species Richness ~ Area_  

```{r, echo=TRUE}
SAR0 <- lm(spec_num ~ area, data = islands)
summary(SAR0)

ggplot(islands, aes(area, spec_num)) +
  geom_point() +
  stat_smooth(method = "lm") +
  stat_poly_eq(method = "lm", formula = y ~ x,
               aes(label = paste(..eq.label.., ..rr.label.., sep = "~~~"))) +
  labs(x = "Area (km^2)", y = "Species richness") +
  theme_bw()

# plot(spec_num ~ area, data = islands)
# car::regLine(SAR0, col = "darkgreen")
```
&nbsp;

## b) Non-linear model  
_Species richness ~ c + Area^z^_  

```{r, echo=TRUE}
SAR1 <- nls(spec_num ~ c + area^z, data = islands,
            start = list(c = 0, z = 1))
summary(SAR1)

ggplot(islands, aes(area, spec_num)) +
  geom_point() +
  geom_smooth(method = "nls",
              formula = y ~ c + x^z,
              method.args = list(start = c(c = 0, z = 1)),
              se = FALSE) +
  labs(title = "Species-area relationship",
       x = "Island area (km^2)", y = "Species number") +
  theme_bw()
```

```{r, echo=TRUE}
# Predictions
newdata <- data.frame(area = seq(min(islands$area), max(islands$area),
                                 length.out = 100))
predicted_richness_SAR1 <- predict(SAR1, newdata = newdata)
head(predicted_richness_SAR1)

# plot(islands$area, islands$spec_num, xlab = "Island area (km)",
#      ylab = "Species number", main = "Species-area relationship")
# points(newdata$area, predicted_richness_SAR1, type = "l",
#        col = "darkred", lwd = 2)
```
&nbsp;

## c) Exercise I  
1. Make histograms of species richness and area  
2. Apply a log10 transformation to species richness and area, make new columns of the transformed variables, and plot histograms of them  
3. Fit a linear model to log predict species richness using log area 
4. Make a plot of the results  

&nbsp;

**Exercise I solutions: **  

_Make histograms of species richness and area_    

```{r, echo=FALSE}
plot_grid(
  ggplot(islands, aes(area)) +
    geom_histogram(color = "black", fill = "grey50") +
    labs(x = "Area (km^2)", y = "Count") +
    theme_bw(),
  ggplot(islands, aes(spec_num)) +
    geom_histogram(color = "black", fill = "grey50") +
    labs(x = "Species richness", y = "Count") +
    theme_bw(),
  nrow = 1)

# par(mfrow = c(1, 2))
# hist(islands$area)
# hist(islands$spec_num)
```
&nbsp;

_Apply a log10 transformation to species richness and area, make new columns of the transformed variables in islands data.frame, and plot histograms of them_  
_&_  
_Fit a linear model to predict species richness using area_  

```{r, echo=FALSE}
islands$log_area <- log10(islands$area + 1)
islands$log_spec_num <- log10(islands$spec_num + 1)

plot_grid(
  ggplot(islands, aes(log_area)) +
    geom_histogram(color = "black", fill = "grey50") +
    labs(x = "Area (km^2) log-transformed", y = "Count") +
    theme_bw(),
  ggplot(islands, aes(log_spec_num)) +
    geom_histogram(color = "black", fill = "grey50") +
    labs(x = "Species richness log-transformed", y = "Count") +
    theme_bw(),
  nrow = 1)

# par(mfrow = c(1, 2))
# hist(islands$log_area)
# hist(islands$log_spec_num)

SAR2 <- lm(log_spec_num ~ log_area, data = islands)
summary(SAR2)

ggplot(islands, aes(log_area, log_spec_num)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, color = "darkgreen") +
  stat_poly_eq(method = "lm", formula = y ~ x,
               aes(label = paste(..eq.label.., ..rr.label.., sep = "~~~"))) +
  labs(x = "Area (km^2) log-transformed",
       y = "Species richness log-transformed") +
  theme_bw()

# par(mfrow = c(1, 1))
# plot(log_spec_num ~ log_area, data = islands)
# car::regLine(SAR2, col = "darkgreen")
```

**Check model assumptions**  

```{r, echo=FALSE}
par(mfrow = c(1, 2))
hist(residuals(SAR0))
hist(residuals(SAR2))
```

&nbsp;

## d) Poisson generalised linear model (GLM)  

```{r, echo=TRUE}
SAR3 <- glm(spec_num ~ log_area, data = islands, family = "poisson")

# 100*(161367-43825)/161367

# Plot Poisson GLM
ggplot(islands, aes(log_area, spec_num)) +
  geom_point() +
  geom_smooth(method = "glm", formula = y ~ x,
              method.args = list(family = "poisson"),
              color = "darkblue") +
  labs(title = "GLM Poisson",
       x = "Area log-transformed (log10 km2)",
       y = "Species richness") +
  theme_bw()

# Plot Poisson GLM in log-log space
newdata_SAR3 <- data.frame(log_area = islands$log_area)
newdata_SAR3$pred <- as.numeric(predict(SAR3,
                                        newdata = newdata_SAR3,
                                        type = "response"))
# Log-transforming the predictions
newdata_SAR3$pred_log10 <- log10(newdata_SAR3$pred + 1)

ggplot(islands, aes(log_area, log_spec_num)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
              color = "darkgreen") +
  geom_line(data = newdata_SAR3, aes(log_area, pred_log10),
            color = "darkblue", size = 1) +
  labs(title = "log-log Species-area relationship",
       x = "Area log-transformed (log10 km2)",
       y = "Species richness log-transformed") +
  theme_bw()
```

In base R, the equivalent would be:  

```{r, echo=TRUE, eval = FALSE}
plot(islands$log_area, islands$log_spec_num,
     xlab = "log10 island area (log10 km)",
     ylab = "log10 species number",
     main = "log-log Species-area relationship")
car::regLine(SAR2)

newdata_SAR3 <- data.frame(log_area = seq(min(islands$log_area),
                                          max(islands$log_area),
                                          length.out = 100))
predicted_richness_SAR3 <- predict(SAR3, newdata = newdata_SAR3,
                                   type = "response")
points(newdata_SAR3$log_area, log10(predicted_richness_SAR3 + 1),
       type = "l", col = "purple", lwd = 2)
```

Now, let's plot all models together.  

```{r}
# Make a data frame with all the predictions
newdata_all <- data.frame(area = islands$area,
                          log_area = islands$log_area)

newdata_all$pred_SAR0 <- as.numeric(predict(SAR0,
                                            newdata = newdata_all,
                                            type = "response"))

newdata_all$pred_SAR1 <- as.numeric(predict(SAR1,
                                            newdata = newdata_all,
                                            type = "response"))

newdata_all$pred_SAR2 <- as.numeric(predict(SAR2,
                                            newdata = newdata_all,
                                            type = "response"))
newdata_all$pred_SAR2 <- (10^newdata_all$pred_SAR2)-1

newdata_all$pred_SAR3 <- as.numeric(predict(SAR3,
                                            newdata = newdata_all,
                                            type = "response"))
# newdata_all$pred_SAR3_exp10 <- (10^newdata_all$pred_SAR3) - 1

# we also need to reverse the log-transformation for area
# newdata_all$logarea_exp10 <- (10^newdata_all$log_area)-1

# Putting the table in a long format
newdata_all <- pivot_longer(newdata_all, cols = contains("SAR"), names_to = "Model", values_to = "Predicted values")

ggplot(newdata_all, aes(area, `Predicted values`)) +
  geom_point(data = islands, aes(area, spec_num)) +
  geom_line(aes(color = Model), size = 1) +
  scale_color_brewer(palette = "Set1") +
  labs(x = "Area (km^2)", y = "Species richness") +
  theme_bw()

```

With base R, the code would be like this:

```{r, echo=TRUE, eval = FALSE}
plot(islands$area, islands$spec_num, xlab = "Island area (km?)",
     ylab = "Species number", main = "Species-area relationship")

# plot the simple linear model
car::regLine(SAR0, col="darkgreen")

# plot the non-linear power law model
points(newdata$area, predicted_richness_SAR1, type = "l",
       col = "darkred", lwd = 2)

# plot the log-log space model in untransformed space
newdata_SAR2 <- data.frame(log_area = log10(seq(min(islands$area),
                                                max(islands$area),
                                                length.out = 100) + 1))
predicted_richness_SAR2 <- predict(SAR2, newdata = newdata_SAR2)
points((10^newdata_SAR2$log_area) - 1, (10^predicted_richness_SAR2) - 1,
       type = "l", col = "purple", lwd = 2)
points(newdata$area, (10^predicted_richness_SAR2) - 1, type = "l",
       col = "darkgreen", lwd = 2, lty = 2)

# plot the Poisson model in untransformed space
predicted_richness_SAR3 <- predict(SAR3, newdata = newdata_SAR3,
                                   type = "response")
points((10^newdata_SAR3$log_area) - 1, predicted_richness_SAR3,
       type = "l", col="darkblue", lwd=2)
```
&nbsp;

**How do these models compare?**  

```{r}
AIC(SAR0, SAR1, SAR2, SAR3)
```

**_Warning_**: You cannot compare AIC values among models with different response variables or different distribution families. This means we can here only compare the AIC values of the first two models!  

&nbsp;

# 3. Multi-predictor model of island biodiversity

## a) Exercise II    

1.  Apply a log10 transformation to area, elevation, precipitation, and temperature following:  
_Kreft, H. et al. 2008. Global diversity of island floras from a macroecological perspective. Ecology Letters 11:116-127._  

2. Simplify the geology column; change all values that are **floor** or **volcanic** to **oceanic** and afterwards set all values that are not **atoll**, **fragment**, **oceanic** or **shelf** to NA.  

3. Fit a multiple linear regression to predict biodiversity patterns on islands using the following biogeographic and environmental
variables: island area, distance, elevation, temperature, precipitation and island geology.  

4. Use manual backward selection to specify the best model  

**Hints**    
1. Remove missing observations from your data  
2. Use 'drop1(mod, test="F")'    
3. Use 'drop1(update(Multi.mod,~ . -log_max_elev), test = "F")'    

&nbsp;

**Solutions to Exercise II**

_Transform your data_
```{r, echo=FALSE}
islands$log_area <- log10(islands$area)
islands$log_spec_num <- log10(islands$spec_num + 1)

islands$log_max_elev <- log10(islands$max_elev + 1)
islands$log_mean_prec <- log10(islands$mean_prec)
islands$log_mean_temp <- log10(islands$mean_temp + 5)
str(islands)
```

&nbsp;

_Simplify the geology column_
```{r,echo=FALSE}
islands$geology <- as.character(islands$geology)
islands$geology[which(islands$geology == "floor" |
                        islands$geology == "volcanic")] <- "oceanic"
islands$geology[which(islands$geology %in%
                        c("floor" , "volcanic"))] <- "oceanic"

islands$geology[which(islands$geology == "floor")] <- "oceanic"
islands$geology[which(islands$geology == "volcanic")] <- "oceanic"


islands$geology[which(!islands$geology %in% c("atoll" , "fragment", "oceanic" , "shelf"))] <- NA

summary(as.factor(islands$geology))
```
&nbsp;

_Fit multiple regression model_  

```{r, echo = FALSE}
colnames(islands)[c(6, 8, 15:19)]
new_islands <- islands[, c(6, 8, 15:19)]
new_islands <- na.omit(new_islands) # remove observations with NAs

Multi.mod <- lm(log_spec_num ~ log_area + dist + log_max_elev +
                  log_mean_temp + log_mean_prec + geology,
                data = new_islands)

summary(Multi.mod)

crPlots(Multi.mod)
```
&nbsp;

_Can we improve predictions of species richness by removing individual model terms?_  

```{r,echo=FALSE}
Multi.mod1 <- update(Multi.mod, ~ . -log_max_elev) 
AIC(Multi.mod, Multi.mod1)
```
&nbsp;

# 4. The General dynamic model of island biogeography (GDM)  

```{r, echo=TRUE}
# Assign archipelago names for 4 major archipelagos
islands$archipelago <- NA
islands$archipelago[which(islands$entity_ID %in% c(169:188))] <- "Hawaii"
islands$archipelago[which(islands$entity_ID %in% c(145:151))] <- "Canaries"
islands$archipelago[which(islands$entity_ID %in% c(154:168))] <- "Galapagos"
islands$archipelago[which(islands$entity_ID %in% c(131:139))] <- "Azores"
islands_gdm <- islands[!is.na(islands$archipelago),]

# Log-transforming age
islands_gdm$log_age_MA <- log10(islands_gdm$age_MA)
```
&nbsp;

**The General dynamic model of island biogeography (GDM) applied: just Hawaii (part I)**

_Species = Area x Time x Time^2^_    

```{r, echo=TRUE}
# subset data
Hawaii <- islands_gdm[which(islands_gdm$archipelago == "Hawaii"), ]

# The ATT^2 model
gdm1 <- lm(spec_num ~ log_area + age_MA + I(age_MA^2), data = Hawaii)
summary(gdm1)
crPlots(gdm1)
```
&nbsp;

**The General dynamic model of island biogeography (GDM) applied: just Hawaii (part II)**

_Species = log(Area) x log(Time) x log(Time)^2^_    

```{r, echo=TRUE}
gdm2 <- lm(spec_num ~ log_area + log_age_MA + I(log_age_MA^2), data = Hawaii)
summary(gdm2)
AIC(gdm1, gdm2)

crPlots(gdm2)
```
&nbsp;

**A linear mixed effects model for all 4 archipelagos**

_Species = log(Area) + log(Time) + log(Time)^2^ + (1|archipelago)_    

```{r, echo=TRUE}
gdm_all <- lme4::glmer(spec_num ~ log_area + log_age_MA + I(log_age_MA^2) +
                         (1|archipelago), data = islands_gdm, family = poisson)
summary(gdm_all)

effect_plot(gdm_all, pred = "log_area", outcome.scale = "response", plot.points = TRUE)
effect_plot(gdm_all, pred = "log_age_MA", outcome.scale = "response", plot.points = TRUE)
model_performance(gdm_all, metrics = "common")
```
