---
title: "Modern concepts and methods in Macroecology and Biogeography"
subtitle: "Day 7: Phylogenetic diversity"
author: "Dylan Craven & Patrick Weigelt"
date: "June 5th, 2019"
output: 
  html_document: 
    fig_height: 7
    theme: flatly
    toc: yes
    toc_depth: 3
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

&nbsp;
_[back to overview](index.html)_
&nbsp;
&nbsp;


**Preparation:** Please open your RStudio project and download the new data ('Phylogeny_ConservativeCon_Checklist.zip', 'palms_specsxsites_phylo.csv' and 'palmtree_pruned.nex') for today from [Stud-IP](https://studip.uni-goettingen.de/dispatch.php/course/files?cid=a16be33e4d2eb3c1c82a459c44495012) and unzip and copy them into your .Rproj folder '/Data/'. You can use `getwd()` to locate your current working directory, which should be your project folder. Please install the following R-packages using `install.packages()`:  

* ape  
* Taxonstand  
* rgdal  
* rgeos  
* maptools  
* plyr  
* psych
* picante  
* pez  

&nbsp;

**Load R packages & island data set**  


```{r message=FALSE, warning=FALSE, include=FALSE}
library(plyr)
library(rgdal)
library(rgeos)
library(maptools)
library(Taxonstand)
library(ape)
library(picante)
library(pez)
library(psych)
library(classInt)
library(colorRamps)
```

```{r eval=FALSE}
library(plyr) # basic data manipulation

library(rgdal) #  Geospatial fun
library(rgeos) #  Geospatial fun
library(maptools)# Geospatial fun

library(Taxonstand) # Taxonomic standardisation of plant species names
library(ape) # Analyses of phylogenetics and evolution
library(picante) # Phylogenies and ecology
library(pez) # Phylogenies and ecology

library(psych) # basic stats stuff
library(classInt) # for plotting nice colors
library(colorRamps) # for plotting nice colors
```
&nbsp;

## 1. Load data into R  

### a. Load a phylogeny \& look at structure of 'multiPhylo' object  


**load Multi tree object into R**  

Data from Faurby, S., Eiserhardt, W.L., Baker, W.J. & Svenning, J.-C. (2016).  
    An all-evidence species-level supertree for the palms (Arecaceae).  
    _Molecular Phylogenetics and Evolution_, 100, 57-69.  

```{r echo=TRUE}
palmtrees <- read.nexus("Data/Phylogeny_ConservativeCon_Checklist.nex")

palmtrees

palmtrees[[1]]

palmtree <- palmtrees[[1]]
str(palmtree)

head(palmtree$tip.label)
length(palmtree$tip.label)

is.rooted(palmtree) # does the phylogeny have a root? i.e. can the most basal ancestor be identified?
```
&nbsp;


*Explanation of structure of a phylogeny in R*  
**edge**: a two-column matrix where each row represents a branch (or edge) of the tree. The nodes and the tips are symbolized with integers.The n tips are numbered from 1 to n, and the m (internal) nodes from n+1 to n+m (the root being n + 1).For each row, the first column gives the ancestor.    

**edge.length (optional)**: A numeric vector giving the lengths of the branches given by edge.  

**tip.label**: A vector of mode character giving the labels of the tips. The order of these labels corresponds to the integers 1 to n in edge.    

**Nnode**: An integer value giving the number of nodes in the tree (m).  

**node.label** (optional): A vector of mode character giving the labels of the nodes (ordered in the same way as the tip.label).    

**root.edge** (optional): A numeric value giving the length of the branch at the root if it exists.  

```{r echo=TRUE, fig.height=9, fig.width=9}

plot(palmtree,type="fan",cex=0.3, edge.color="gray70",tip.color="#ef8a62")

```
&nbsp;


### b.  Loading palm distribution data  

Data from  Kreft, H., Sommer, J.H. & Barthlott, W. (2006).  
    The significance of geographic range size for spatial diversity  
    patterns in Neotropical palms. _Ecography_, 29, 21-30.  


```{r echo=TRUE}
species <- read.csv("Data/palms_species_per_gridcell.csv",sep=",", stringsAsFactors = FALSE)
head(species)
length(unique(species$grid_id)) # number of grid cells
```
&nbsp;


### c. Exercise I  

**Tasks**:    
1. Add a column for the full species name with "_" as separator: "species".      
2. How many palm species occur in the Americas?  
3. How many and which species in the palm distribution dataset are missing from the phylogeny?  
&nbsp;

**Exercise I solutions:**  
_Add a column for the full species name with an underscore as the separator: "species"_  
```{r echo=FALSE}
species$species <- paste(species$GENUS,species$EPITHET, sep="_")
species$species <- paste(species$GENUS,"_",species$EPITHET, sep="")
unique(species$species)[1:5]
```
&nbsp;

_How many palm species occur in the Americas?_

```{r echo=FALSE}
length(unique(species$species))
```
&nbsp;

_How many and which species in the palm distribution dataset are missing from the phylogeny?_

```{r echo=FALSE}
length(unique(species$species[which(!species$species %in% palmtree$tip.label)]))

# first five palm species
unique(species$species[which(!species$species %in% palmtree$tip.label)])[1:5]
```
&nbsp;



## 2. Taxonomic standardization  

**Write species missing from phylogeny into a vector**
```{r echo=TRUE}

specmissing <- unique(species$species[which(!species$species %in% palmtree$tip.label)])

specmissing <- gsub("_"," ",specmissing) # # replace underscore with a space
```
&nbsp;


**Match palm species names to those in [The Plant List](http://www.theplantlist.org/)**

```{r echo=TRUE, message=FALSE, warning=FALSE}
taxstand <- TPL(specmissing, diffchar = 2, max.distance = 1)
head(taxstand)[,11:16]
```
&nbsp;

**Replace old names by new names \& overwrite the combined species name**
```{r echo=TRUE}
for (i in 1:nrow(taxstand)){
  species$GENUS[which(species$GENUS == taxstand$Genus[i] & species$EPITHET == taxstand$Species[i])] <- taxstand$New.Genus[i]
  species$EPITHET[which(species$GENUS == taxstand$Genus[i] & species$EPITHET == taxstand$Species[i])] <- taxstand$New.Species[i]
}

species$species <- paste(species$GENUS,species$EPITHET, sep="_") # overwrite the combined species name
```
&nbsp;

**Check names that are not in the phylogeny**  
```{r echo=TRUE}
specmissing <- unique(species$species[which(!species$species %in% palmtree$tip.label)])
specmissing
length(specmissing)
species <- species[which(!species$species %in% specmissing),]
```
&nbsp; 

_Missing species would have to be looked up and added manually to the phylogeny by an expert._   
_If missing species cannot be added reliably, one can remove them from the dataset for analysis._  


## 3. Loading grid shapefile and changing resolution  


**Read in polygon shapefile**  
```{r echo=TRUE, message=FALSE, warning=FALSE}
americas <- readOGR("Data/americas.shp")
grid <- readOGR("Data/30min_grid_select50%.shp", integer64="allow.loss")
names(grid@data)[1] <- "grid_id"
```
&nbsp; 

**Convert the grid shapefile to a coarser resolution to have fewer grid cells**  

```{r echo=TRUE}
grid_centroids <- gCentroid(grid, byid=TRUE) # make spatial points dataframe from grid
grid_centroids <- SpatialPointsDataFrame(grid_centroids, data = data.frame(grid_centroids@coords))
names(grid_centroids@data) <- c("longitude","latitude")
par(mfrow=c(1,1))
plot(grid_centroids)
plot(americas, add=TRUE)
```
&nbsp; 

**Add coordinates to SpatialPolygonsDataframe**  

```{r echo=TRUE}
grid@data <- cbind(grid@data,grid_centroids@data)
head(grid@data)
```
&nbsp; 

**Make new IDs for neighbouring cells**  
```{r echo=TRUE}
range(grid@data$longitude)
range(grid@data$latitude)

longitude <- seq(-116,-34, by=2)
latitude  <- seq(-35,35, by=2)

new_ID_matrix <- matrix(c(1:(length(longitude)*length(latitude))), nrow=length(longitude), ncol=length(latitude) , dimnames = list(longitude, latitude), byrow=TRUE)

grid@data$new_ID <- NA

for(i in 1:nrow(grid@data)){
  grid@data$new_ID[i] <-   new_ID_matrix[which(longitude < grid@data$longitude[i] & (longitude + 2) > grid@data$longitude[i]), which(latitude < grid@data$latitude[i] & (latitude + 2) > grid@data$latitude[i])]
}

head(grid@data)
```
&nbsp; 

**Join neighbouring cells based on new IDs \& aggregate species distribution data using the new grid**  
```{r echo=TRUE}
grid_2degrees <- unionSpatialPolygons(grid, grid@data$new_ID)
uniqueIDs <- unique(grid@data$new_ID)
grid_2degrees <- SpatialPolygonsDataFrame(grid_2degrees, data=data.frame(new_ID=uniqueIDs, row.names = uniqueIDs))
head(grid_2degrees@data)
plot(grid_2degrees)
plot(americas, add=TRUE)

species <- join(species, grid@data, by="grid_id", type="inner")
head(species)
```
&nbsp;

## 4. Calculating phylogenetic community metrics  


**Prune phylogeny to exclude species not in palm distribution data**  
_Make sure that the same number of species in your data set are in your phylogeny_

```{r echo=TRUE}
palmtree_pruned <- drop.tip(palmtree,palmtree$tip.label[which(!palmtree$tip.label %in% unique(species$species))])
length(palmtree_pruned$tip.label) # 
length(unique(species$species))

write.nexus(palmtree_pruned, file="Data/palmtree_pruned.nex")
```
&nbsp;

**Convert long table to species by grid-cell table (a community matrix)**
```{r echo=TRUE}
species <- unique(species[,c("new_ID","species")])
species_grid <- as.data.frame.matrix(table(species))
species_grid<-as.matrix(species_grid)

write.csv(species_grid, file = "Data/palms_specsxsites_phylo.csv")
```
&nbsp;

**Calculate phylogenetic diversity**  

```{r message=FALSE, warning=FALSE, r,echo=TRUE}
palmtree_pruned<-read.nexus("Data/palmtree_pruned.nex") # if you have a hard time reading in the large palm phylogeny
species_grid <- as.matrix(read.csv("Data/palms_specsxsites_phylo.csv", row.names = 1)) # in case you got stuck above

pd_palms <- pd(species_grid, palmtree_pruned)
head(pd_palms)
```
&nbsp;

`pd()` calculates Faith's PD and a corrected version of it being the residuals of a regression with total abundance or species richness per plot. Lets look into this:

```{r message=FALSE, warning=FALSE, echo=TRUE, fig.height=5, fig.width=8}
pd_model <- lm(PD ~ SR, data = pd_palms)
pd_palms$residuals <- residuals(pd_model)
par(mfrow=c(1,2))
plot(pd_palms$SR, pd_palms$PD, main="Species richness vs Faith's PD", xlab="Species number",  ylab="Faith's PD", cex.main=0.8,cex.axis=0.8)
abline(pd_model)
plot(pd_palms$SR, pd_palms$residuals, main="Species richness vs Faith's PD residuals", xlab="Species number",  ylab="Faith's PD residuals", cex.main=0.8,cex.axis=0.8)
```


**Calculate generic phylogenetic metrics (absolute and standardized)**  

Phylogenetic diversity metrics are standardized by first using a null model to generate expected values.  
Then, the observed value is compared to the null, usually using standardized effect sizes:  
$SES = (observed - mean.null)/sd.null$, where mean.null is the mean of the null values and sd.null is the standard deviation of the null values.  


```{r message=FALSE, warning=FALSE, echo=TRUE, fig.height=5, fig.width=8}
c.data <- comparative.comm(palmtree_pruned , species_grid)
pd.null<-generic.null(c.data, c(.pd,.mpd,.mntd),null.model = "richness",comp.fun=.ses,permute=100) 
colnames(pd.null) <- c("Faith_PD","Corrected_FaithPD","MPD", "MNTD")
rownames(pd.null) <- sites(c.data)
pd.null<-data.frame(pd.null)
pd.null$new_ID<-as.numeric(rownames(pd.null))

pd.out    <- pd.null[,c(21, 1,13,3,15,4,16)]  # select columns for final data set
pd.out [is.na(pd.out )] <- 0  # change NAs to zeroes; no phylogenetic diversity if grid has only 1 spp!

par(mfrow=c(1,2))
plot(pd_palms$SR, pd_palms$residuals, main="Species richness vs Faith's PD residuals", xlab="Species number",  ylab="Faith's PD residuals", cex.main=0.8,cex.axis=0.8)
plot(pd.out$Faith_PD.observed, pd.out$Faith_PD.SES,main="Faith's PD vs standardised Faith's PD",cex.main=0.8,xlab="observed Faith's PD",ylab="standardised Faith's PD",cex.axis=0.8)
```
&nbsp;


### a. Exercise II  

**Tasks**:  
1. Create a data.frame called 'palmdiversity' with species richness, PD and residual PD from 'pd_palms' and standardized pd metrics from 'pd.out'.    
2. Plot correlations among the diversity metrics.   
3. Join the 'palmdiversity' data.frame to the grid_2degrees shapefile to be able to link the diversity metrics and the poylgons for plotting.   
4. Plot maps with colours according to the diversity metrics.  

**Exercise II solutions:**

_Create a data.frame called 'palmdiversity' by joining 'pd_palms' with 'pd.out'_

```{r echo=FALSE}
palmdiversity<- data.frame(new_ID = as.numeric(rownames(species_grid)), pd_palms) 
palmdiversity<-join(palmdiversity,pd.out, by="new_ID")
head(palmdiversity)
```
&nbsp;

_Plot correlations among the diversity metrics_  
```{r echo=FALSE, fig.height=10, fig.width=10}
pairs.panels(palmdiversity[,c(3:10)], method="pearson",density=F,ellipses=F, hist.col = "white")
```
&nbsp;

_Join the palmdiversity data.frame to the shapefile with the new gridcell IDs_  

```{r echo=FALSE}

grid_2degrees@data <- join(grid_2degrees@data, palmdiversity, by = "new_ID", type="left")
head(grid_2degrees@data)
```
&nbsp;


_Plot maps with different colours for each diversity metric_  

*Species richness, PD, MPD \& MNTD (absolute)*  
```{r echo=FALSE, fig.height=10, fig.width=6.6, message=FALSE, warning=FALSE}
par(mfrow=c(2,2),mar=c(1,1,1,1), oma=c(0,0,0,0))

my.class.fr<-classIntervals(grid_2degrees@data$SR,n=10,style="equal") # bin data into n quantiles
my.pal<-matlab.like(10)
my.col.fr<-findColours(my.class.fr,my.pal) # ramp colors based on classInts

plot(grid_2degrees, col=my.col.fr,border=NA, axes=FALSE)
title(main="Species_richness",line=-3)
plot(americas, add=TRUE)

my.class.fr<-classIntervals(grid_2degrees@data$Faith_PD.observed,n=10,style="equal") # bin data into n quantiles
my.pal<-matlab.like(10)
my.col.fr<-findColours(my.class.fr,my.pal) # ramp colors based on classInts

plot(grid_2degrees, col=my.col.fr,border=NA, axes=FALSE)
title(main="Faith's PD",line=-3)
plot(americas, add=TRUE)

my.class.mpd<-classIntervals(grid_2degrees@data$MPD.observed,n=10,style="equal") # bin data into n quantiles
my.pal2<-matlab.like(10)
my.col.mpd<-findColours(my.class.mpd,my.pal2) # ramp colors based on classInts

plot(grid_2degrees, col=my.col.mpd, border=NA, axes=FALSE)
title(main="MPD", line = -3)
plot(americas, add=TRUE)

my.class.mntd<-classIntervals(grid_2degrees@data$MNTD.observed,n=10,style="equal") # bin data into n quantiles
my.pal3<-matlab.like(10)
my.col.mntd<-findColours(my.class.mntd,my.pal3) # ramp colors based on classInts

plot(grid_2degrees, col=my.col.mntd, border=NA, axes=FALSE)
title(main="MNTD", line = -3)
plot(americas, add=TRUE)
```
&nbsp;


*PD, MPD \& MNTD (standardized)*  
```{r echo=FALSE, fig.height=5, fig.width=10}
par(mfrow=c(1,3),mar=c(1,1,1,1), oma=c(0,0,0,0))

my.class.fr2<-classIntervals(grid_2degrees@data$Faith_PD.SES,n=10,style="equal") 
my.pal4<-matlab.like(10)
my.col.fr2<-findColours(my.class.fr2,my.pal4) # ramp colors based on classInts

plot(grid_2degrees, col=my.col.fr2, border=NA)
title(main="standardised Faith's PD", line = -3)
plot(americas, add=TRUE)

my.class.fr3<-classIntervals(grid_2degrees@data$MPD.SES,n=10,style="equal") # bin data into n quantiles
my.pal5<-matlab.like(10)
my.col.fr3<-findColours(my.class.fr3,my.pal5) # ramp colors based on classInts

plot(grid_2degrees, col=my.col.fr3, border=NA)
title(main="standardised MPD", line = -3)
plot(americas, add=TRUE)

my.class.fr4<-classIntervals(grid_2degrees@data$MNTD.SES,n=10,style="equal") # bin data into n quantiles
my.pal6<-matlab.like(10)
my.col.fr4<-findColours(my.class.fr4,my.pal6) # ramp colors based on classInts

plot(grid_2degrees, col=my.col.fr4, border=NA)
title(main="standardised MNTD", line = -3)
plot(americas, add=TRUE)
```
&nbsp;